---
title: "Real-World QC and Imprinting Analysis: GSE110697 Dataset"
subtitle: "Run methylation-array QC and imprintomeR analysis on a public EPIC Wilms tumor cohort"
author: "imprintomeR contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 10
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{GSE110697 QC and Imprinting Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "GSE110697_",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This vignette demonstrates a complete imprintomeR workflow on **GSE110697**, a public DNA methylation dataset comparing Wilms tumor primary samples and patient-derived xenografts. You will:

1. **Download** GEO metadata and processed beta values
2. **Prepare** raw IDAT metadata for the MethQcSet QC workflow
3. **Run** methylation-array QC when IDAT files are available
4. **Create** an ImprintomeSet object from QC-clean or processed data
5. **Analyze** imprinting status and visualize results
6. **Export** QC and imprintomeR outputs

GSE110697 uses the Illumina Infinium HumanMethylationEPIC BeadChip (850K). The GEO record includes 78 samples, processed data in the sample table, and a large raw IDAT archive. The processed-beta route keeps the vignette lightweight; the raw-IDAT route shows the full QC workflow and is intended for local execution after downloading the supplementary tar file.

---

## Phase 1: Setup & GEO Download

### Load libraries

```{r load-libraries, message = FALSE, warning = FALSE}
library(imprintomeR)
library(GEOquery)
library(tidyverse)
library(ComplexHeatmap)
library(ggplot2)
library(knitr)

set.seed(2026)
```

### Download GSE110697 metadata and processed values

```{r download-gse110697, eval = TRUE, cache = TRUE}
geo_gse <- "GSE110697"
message("Downloading ", geo_gse, " from GEO...")

tryCatch({
  gse <- getGEO(geo_gse, GSEMatrix = TRUE, AnnotGPL = FALSE, getGPL = FALSE)
  eset <- gse[[1]]

  message("Download successful")
  message("  Samples: ", ncol(eset))
  message("  Features: ", nrow(eset))
}, error = function(e) {
  stop("Failed to download ", geo_gse, ":\n", conditionMessage(e),
       "\nCheck internet connection and GEO availability.")
})
```

### Extract metadata

```{r extract-metadata}
pdata <- pData(eset)

message("Available metadata columns:")
print(colnames(pdata))

message("\nSample overview:")
display_cols <- intersect(
  c("geo_accession", "title", "source_name_ch1", "characteristics_ch1"),
  colnames(pdata)
)
head(pdata[, display_cols, drop = FALSE], n = 6)
```

### Build sample metadata

```{r build-meta}
sample_title <- if ("title" %in% colnames(pdata)) pdata$title else pdata$geo_accession

meta <- data.frame(
  Sample_Name = pdata$geo_accession,
  GEO_Accession = pdata$geo_accession,
  Sample_Title = sample_title,
  Sample_Group = dplyr::case_when(
    grepl("^APT", sample_title, ignore.case = TRUE) ~ "APT",
    grepl("^AKT", sample_title, ignore.case = TRUE) ~ "AKT",
    TRUE ~ "Other"
  ),
  Platform = "EPIC",
  stringsAsFactors = FALSE
)

# Keep GEO columns that are often useful during interpretation.
optional_cols <- intersect(
  c("source_name_ch1", "characteristics_ch1", "characteristics_ch1.1",
    "characteristics_ch1.2", "characteristics_ch1.3"),
  colnames(pdata)
)
meta <- cbind(meta, pdata[, optional_cols, drop = FALSE])

message("Metadata created:")
message("  Samples: ", nrow(meta))
print(table(meta$Sample_Group))
head(meta, n = 6)
```

---

## Phase 2: Raw IDAT QC Workflow

The raw IDAT archive for GSE110697 is large, so this section is set to `eval = FALSE`. Run it locally when you want full preprocessing and QC from raw array files.

### Download and unpack supplementary IDAT files

```{r download-raw-idat, eval = FALSE}
raw_dir <- file.path("GSE110697_raw")
dir.create(raw_dir, showWarnings = FALSE, recursive = TRUE)

# Downloads GSE110697_RAW.tar from GEO.
supp <- getGEOSuppFiles(geo_gse, makeDirectory = TRUE, baseDir = raw_dir)
print(supp)

tar_file <- rownames(supp)[grepl("RAW\\.tar$", rownames(supp))]
untar(tar_file, exdir = file.path(raw_dir, geo_gse))

# GEO IDAT archives are commonly gzipped inside the tar.
gz_idats <- list.files(
  file.path(raw_dir, geo_gse),
  pattern = "\\.idat\\.gz$",
  recursive = TRUE,
  full.names = TRUE
)
if (length(gz_idats) > 0 && requireNamespace("R.utils", quietly = TRUE)) {
  lapply(gz_idats, R.utils::gunzip, overwrite = TRUE, remove = FALSE)
}
```

### Create QC-ready metadata with Basename paths

`runMethQC()` expects a `Basename` column containing the IDAT path prefix without `_Red.idat` or `_Grn.idat`.

```{r make-idat-meta, eval = FALSE}
idat_files <- list.files(
  file.path(raw_dir, geo_gse),
  pattern = "_(Red|Grn)\\.idat(\\.gz)?$",
  recursive = TRUE,
  full.names = TRUE
)

idat_prefixes <- unique(sub("_(Red|Grn)\\.idat(\\.gz)?$", "", idat_files))
idat_map <- data.frame(
  Basename = idat_prefixes,
  Sample_Title = basename(idat_prefixes),
  stringsAsFactors = FALSE
)

# Match conservatively by sample title. If your unpacked files include GSM IDs
# instead, replace this join key with GEO_Accession.
meta_qc <- meta %>%
  left_join(idat_map, by = "Sample_Title") %>%
  mutate(
    Sample_Name = Sample_Title,
    Sample_Group = Sample_Group,
    Platform = "EPIC"
  )

missing_idat <- meta_qc %>% filter(is.na(Basename))
if (nrow(missing_idat) > 0) {
  warning("Samples without matched IDAT prefixes: ", nrow(missing_idat))
  print(missing_idat[, c("Sample_Name", "Sample_Title", "GEO_Accession")])
}

meta_qc <- meta_qc %>% filter(!is.na(Basename))
head(meta_qc[, c("Sample_Name", "Sample_Group", "Platform", "Basename")])
```

### Run MethQcSet QC

```{r run-meth-qc, eval = FALSE, cache = TRUE}
qcset <- runMethQC(
  meta_qc,
  platform = "EPIC",
  pcutoff = 0.05
)

qc_summary <- qc_summarize(qcset)
str(qc_summary)

qc_matrix <- qc_tables(qcset)[["QC_matrix"]]
message("QC status:")
print(table(qc_matrix$Final.QC))

head(qc_matrix)
```

### QC visualizations

```{r qc-plots, eval = FALSE, fig.width = 10, fig.height = 5}
p_qc_bar <- plot(qcset, type = "qc_bar", outFile = NULL)
print(p_qc_bar)

p_qc_intensity <- plot(qcset, type = "intensity", icutoff = 11, outFile = NULL)
print(p_qc_intensity)

p_qc_detection <- plot(qcset, type = "detection_pval", pcutoff = 0.05, outFile = NULL)
print(p_qc_detection)

p_qc_coverage <- plot(qcset, type = "probe_coverage", outFile = NULL)
print(p_qc_coverage)

p_qc_sex <- plot(qcset, type = "predicted_sex", outFile = NULL)
print(p_qc_sex)
```

### Export QC outputs

```{r export-qc, eval = FALSE}
qc_outdir <- "GSE110697_qc_output"
dir.create(qc_outdir, showWarnings = FALSE, recursive = TRUE)

export(
  qcset,
  outdir = qc_outdir,
  format = c("xlsx", "rds", "txt"),
  prefix = "GSE110697"
)

message("QC outputs exported to: ", qc_outdir)
```

---

## Phase 3: Prepare Beta Values for ImprintomeSet

### Use QC-clean beta values when available

If you ran the raw-IDAT QC workflow above, use PASS samples from `qcset`.

```{r beta-from-qc, eval = FALSE}
qc_matrix <- qc_tables(qcset)[["QC_matrix"]]
pass_samples <- qc_matrix$Sample_Name[qc_matrix$Final.QC == "PASS"]

meta_pass <- meta(qcset)[match(pass_samples, meta(qcset)$Sample_Name), , drop = FALSE]
beta_pass <- beta(qcset)[, pass_samples, drop = FALSE]
det_pval_pass <- if (!is.null(detection_pval(qcset))) {
  detection_pval(qcset)[, pass_samples, drop = FALSE]
} else {
  NULL
}
qc_tables_pass <- lapply(qc_tables(qcset), function(tbl) {
  if (is.data.frame(tbl) && "Sample_Name" %in% colnames(tbl)) {
    tbl[tbl$Sample_Name %in% pass_samples, , drop = FALSE]
  } else {
    tbl
  }
})

qcset_pass <- MethQcSet(
  meta = meta_pass,
  platform = platform(qcset),
  beta = beta_pass,
  detection_pval = det_pval_pass,
  qc_tables = qc_tables_pass,
  aggregation_status = aggregation_status(qcset),
  qc_params = qc_params(qcset)
)

x <- as.ImprintomeSet(
  qcset_pass,
  probeset = readRDS(system.file("extdata", "probesets_hg19.rds", package = "imprintomeR"))[["selected"]],
  genome = "hg19"
)
```

### Processed-beta route from GEO

Use this route when you want to run the imprinting workflow directly from GEO processed beta values.

```{r extract-beta}
beta_matrix <- exprs(eset)

if (nrow(beta_matrix) == 0L) {
  stop("No processed beta matrix was found in the GEO Series Matrix object. ",
       "Run the raw IDAT QC workflow or inspect the GEO sample tables manually.")
}

storage.mode(beta_matrix) <- "numeric"
beta_matrix <- beta_matrix[rowSums(!is.na(beta_matrix)) > 0, , drop = FALSE]

message("Beta matrix dimensions: ", nrow(beta_matrix), " probes x ",
        ncol(beta_matrix), " samples")
message("Beta range: [", round(min(beta_matrix, na.rm = TRUE), 3), ", ",
        round(max(beta_matrix, na.rm = TRUE), 3), "]")

# Align GEO accession sample IDs between beta and metadata.
meta_processed <- meta %>%
  filter(Sample_Name %in% colnames(beta_matrix)) %>%
  arrange(match(Sample_Name, colnames(beta_matrix)))
beta_matrix <- beta_matrix[, meta_processed$Sample_Name, drop = FALSE]

stopifnot(identical(colnames(beta_matrix), meta_processed$Sample_Name))
```

### Validate processed data

```{r validate-data}
pct_missing <- colSums(is.na(beta_matrix)) / nrow(beta_matrix) * 100
message("Missing beta values per sample:")
print(head(sort(pct_missing, decreasing = TRUE), n = 10))

message("Sample names aligned: ", identical(colnames(beta_matrix), meta_processed$Sample_Name))
print(table(meta_processed$Sample_Group))
```

---

## Phase 4: Create ImprintomeSet

### Load data using imprintomeR utilities

```{r load-metabeta}
beta_df <- as.data.frame(beta_matrix)
beta_df <- cbind(NAME = rownames(beta_df), beta_df)

input <- LoadMetaBeta(meta_processed, beta_df, probeset = NULL)

message("Loaded data:")
message("  Samples: ", ncol(input$beta))
message("  Probes:  ", nrow(input$beta))
message("  Metadata columns: ", paste(colnames(input$meta), collapse = ", "))
```

### Create the analysis-ready object

```{r create-imprintomeset}
probesets <- readRDS(
  system.file("extdata", "probesets_hg19.rds", package = "imprintomeR")
)

x <- ImprintomeSet(
  beta = input$beta,
  meta = input$meta,
  probeset = probesets[["selected"]],
  genome = "hg19",
  assay = "EPIC"
)

message("ImprintomeSet created:")
message("  Samples: ", ncol(beta(x)))
message("  Probes:  ", nrow(beta(x)))
message("  Genome:  ", genome(x))
message("  Assay:   ", assay(x))
print(summary(x))
```

---

## Phase 5: Run Workflow & Compute Statistics

### Execute imprintomeR analysis

```{r run-imprintome}
message("Running imprintome analysis...")
x <- runImprintome(x, probeset = "selected", ids_cutoff = 0.2)
message("Analysis complete")

res <- results(x)
message("Results available:")
print(names(res))

status <- res[["AnalyzeImprintStatus.selected"]]
head(status, n = 5)
```

### Compute per-group statistics

```{r compute-stats}
message("Imprinting status by group:")
print(table(status[, c("Sample_Group", "Mechanism")]))

stat_summary <- status %>%
  group_by(Sample_Group, Mechanism) %>%
  summarize(
    n = n(),
    median_IDS = median(IDS, na.rm = TRUE),
    median_Angle = median(Angle, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Sample_Group, desc(n))

stat_summary
```

### Review key metrics

```{r review-metrics}
message("IDS distribution:")
print(summary(status$IDS))

message("\nMechanism classification:")
print(table(status$Mechanism))

message("\nAngle distribution:")
print(summary(status$Angle))
```

---

## Phase 6: Visualizations

```{r choose-sample}
sample_id_1 <- colnames(beta(x))[2]
message("Representative sample for single-sample plots: ", sample_id_1)
```

### Plot 1: Polar scatter

```{r plot-polar, fig.width = 8, fig.height = 8}
p_polar <- plot(
  x,
  plot_type = "polar",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "Sample_Group",
  outFile = NULL
)
print(p_polar)
```

### Plot 2: Beeswarm by origin

```{r plot-beeswarm-origin, fig.width = 12, fig.height = 6}
p_beeswarm_origin <- plot(
  x,
  plot_type = "beeswarm_origin",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  outFile = NULL
)
print(p_beeswarm_origin)
```

### Plot 3: Mirror density

```{r plot-mirror-density, fig.width = 10, fig.height = 6}
p_mirror <- plot(
  x,
  plot_type = "mirror_density",
  probeset = "selected",
  outFile = NULL
)
print(p_mirror)
```

### Plot 4: Circular heatmap

```{r plot-circular, fig.width = 10, fig.height = 10}
p_circular <- plot(
  x,
  plot_type = "circular_heatmap",
  probeset = "selected",
  outFile = NULL
)
p_circular
```

### Plot 5: Beta heatmap by probe

```{r plot-heatmap-by-probe, fig.width = 10, fig.height = 8}
p_heatmap <- plot(
  x,
  plot_type = "heatmap_by_probe",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  annoColumn = "Sample_Group",
  outFile = NULL
)
draw(p_heatmap, merge_legend = TRUE)
```

### Plot 6: Beta heatmap by gene

```{r plot-heatmap-by-gene, fig.width = 10, fig.height = 12}
p_heatmap_by_gene <- plot(
  x,
  plot_type = "heatmap_by_gene",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  annoColumn = "Sample_Group",
  clusterRows = TRUE,
  clusterColumns = TRUE,
  outFile = NULL
)
print(p_heatmap_by_gene)
```

### Plot 7: Radar plot

```{r plot-radar, fig.width = 8, fig.height = 8}
p_radar <- plot(
  x,
  plot_type = "radar",
  sample_id = sample_id_1,
  probeset = "selected",
  outFile = NULL
)
print(p_radar)
```

### Plot 8: Beeswarm by chromosome

```{r plot-beeswarm-chr, fig.width = 12, fig.height = 6}
p_beeswarm_chr <- plot(
  x,
  plot_type = "beeswarm_chr",
  sample_id = sample_id_1,
  probeset = "selected",
  outFile = NULL
)
print(p_beeswarm_chr)
```

### Plot 9: Rainfall plot

```{r plot-rainfall, fig.width = 12, fig.height = 6}
p_rainfall <- plot(
  x,
  plot_type = "rainfall",
  sample_id = sample_id_1,
  probeset = "classifier3",
  outFile = NULL
)
print(p_rainfall)
```

### Store generated plots

```{r store-plots}
plot_list <- plots(x)
new_plots <- list(
  "polar.selected" = p_polar,
  "beeswarm_origin.selected" = p_beeswarm_origin,
  "mirror_density.selected" = p_mirror,
  "circular.selected" = p_circular,
  "heatmap.selected" = p_heatmap,
  "heatmap_by_gene.selected" = p_heatmap_by_gene,
  "radar.sample1" = p_radar,
  "beeswarm_chr.sample1" = p_beeswarm_chr,
  "rainfall.sample1" = p_rainfall
)
new_plots <- new_plots[!vapply(new_plots, is.null, logical(1))]
plot_list[names(new_plots)] <- new_plots
plots(x) <- plot_list
names(plots(x))
```

---

## Phase 7: Results Interpretation

### Summary statistics

```{r interpret-summary}
normal_n <- sum(status$Mechanism == "NORMAL", na.rm = TRUE)
total_n <- nrow(status)
normal_pct <- normal_n / total_n * 100
abnormal_pct <- 100 - normal_pct

message("=== OVERALL IMPRINTING SUMMARY ===")
message(sprintf("Normal imprinting:   %6.1f%% (%d/%d samples)",
                normal_pct, normal_n, total_n))
message(sprintf("Altered imprinting:  %6.1f%% (%d/%d samples)",
                abnormal_pct, total_n - normal_n, total_n))

message("\n=== BY SAMPLE GROUP ===")
for (group in unique(status$Sample_Group)) {
  idx <- status$Sample_Group == group
  n <- sum(idx)
  normal <- sum(status$Mechanism[idx] == "NORMAL", na.rm = TRUE)
  message(sprintf("%12s: %d samples, %d normal (%.1f%%)",
                  group, n, normal, 100 * normal / n))
}
```

### Identify samples with aberrant imprinting

```{r identify-abnormal}
abnormal <- status %>% filter(Mechanism != "NORMAL")
sample_col <- intersect(c("Sample_Name", "Sample_ID"), colnames(status))[1]

if (nrow(abnormal) > 0) {
  message("=== ABERRANT IMPRINTING SAMPLES ===")
  message("Sample count: ", nrow(abnormal))
  print(abnormal %>% select(all_of(sample_col), Sample_Group, Mechanism, IDS, Angle) %>% head(n = 12))
} else {
  message("All samples show normal imprinting patterns by the selected IDS cutoff.")
}
```

### Pair-aware comparison

APT and AKT sample titles use shared numeric suffixes for many primary/xenograft pairs. This block creates a pair ID and summarizes imprinting scores across paired labels when both are available.

```{r pair-aware-summary}
status_pairs <- status %>%
  mutate(
    Sample_Title = meta(x)$Sample_Title[match(.data[[sample_col]], meta(x)$Sample_Name)],
    Pair_ID = sub("^[A-Za-z]+", "", Sample_Title)
  )

paired_ids <- status_pairs %>%
  count(Pair_ID) %>%
  filter(n > 1) %>%
  pull(Pair_ID)

pair_summary <- status_pairs %>%
  filter(Pair_ID %in% paired_ids) %>%
  select(Pair_ID, all_of(sample_col), Sample_Title, Sample_Group, Mechanism, IDS, Angle) %>%
  arrange(Pair_ID, Sample_Group)

head(pair_summary, n = 12)
```

### Key findings template

```{r key-findings}
message("=== KEY BIOLOGICAL INSIGHTS TEMPLATE ===")
message("1. Dataset: GSE110697, Wilms tumor primary and xenograft methylation cohort")
message("2. Platform: Illumina HumanMethylationEPIC / 850K")
message("3. Samples analyzed: ", nrow(status))
message("4. IDS range: [",
        round(min(status$IDS, na.rm = TRUE), 3), ", ",
        round(max(status$IDS, na.rm = TRUE), 3), "]")
message("5. Most common mechanism: ",
        names(sort(table(status$Mechanism), decreasing = TRUE))[1])
message("6. Use pair_summary to inspect primary/xenograft concordance by matched suffix.")
```

---

## Phase 8: Export Results

```{r export-results, eval = FALSE}
manifest <- export(
  x,
  outdir = "GSE110697_imprintome_output",
  result_names = "AnalyzeImprintStatus.selected",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

head(manifest)
```

`export()` writes `results(x)` and optionally `plots(x)`. Save extra summary tables only when you need custom reporting beyond the export manifest.

---

## Session Information & Reproducibility

```{r session-info}
message("=== REPRODUCIBILITY ===")
message("R version: ", R.version$version.string)
message("imprintomeR version: ", packageVersion("imprintomeR"))
message("Execution date: ", Sys.Date())
message("Random seed: 2026")

sessionInfo()
```

---

## Jupyter Notebook Conversion Guide

This vignette is designed for conversion to Jupyter notebooks. To convert:

```bash
# 1. Render this Rmd to HTML (in R):
rmarkdown::render("vignettes/GSE110697.Rmd")

# 2. Convert HTML to Jupyter notebook:
jupyter nbconvert --to notebook GSE110697.html --output GSE110697.ipynb

# 3. Or use Jupytext to convert Rmd directly:
jupytext --to notebook vignettes/GSE110697.Rmd
```

**Tips for Jupyter users:**
- The raw-IDAT QC chunks are disabled by default; change `eval = FALSE` to `eval = TRUE` after downloading the raw archive.
- Use `cache = TRUE` on GEO and QC chunks to avoid repeated downloads and IDAT reads.
- Set `outFile` in plot calls to save publication-ready PDFs or PNGs.
- Keep QC and imprintome outputs in separate directories for easier review.

---

## References

- **GSE110697**: [NCBI GEO record](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE110697)
- **GSE110699**: SuperSeries for primary tumor and xenograft comparisons
- **imprintomeR documentation**: See package vignettes
- **GEOquery**: Bioconductor package for GEO data access
- **MethQcSet**: QC container for raw methylation array preprocessing
- **ImprintomeSet**: Analysis container for imprinting studies

---

## License & Attribution

This vignette uses publicly available data from NCBI Gene Expression Omnibus (GEO). Please cite the original GSE110697 study and imprintomeR when using this analysis.
