---
title: "Real-World Imprinting Analysis: GSE52576 Dataset"
subtitle: "Download, preprocess, and analyze methylation data from a public GEO dataset"
author: "imprintomeR contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 10
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{GSE52576 Real-World Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "GSE52576_",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This vignette demonstrates imprintomeR on **GSE52576**, a real-world methylation dataset from the Gene Expression Omnibus (GEO). You will:

1. **Download** normalized beta values from GEO using GEOquery
2. **Create** an ImprintomeSet analysis-ready object
3. **Analyze** imprinting status across samples
4. **Visualize** results with all available plot types
5. **Interpret** biological patterns and aberrant imprinting

The workflow is optimized for **reproducibility** and **Jupyter notebook conversion**.

---

## Phase 1: Setup & Data Download

### Load libraries

```{r load-libraries, message = FALSE, warning = FALSE}
library(imprintomeR)
library(GEOquery)
library(tidyverse)
library(ComplexHeatmap)
library(ggplot2)
library(knitr)

# For reproducibility
set.seed(2026)
```

### Download GSE52576 from GEO

GSE52576 is a publicly available dataset of	Genome-wide parent-of-origin DNA methylation analysis using Illumina 450K methylation arrays.

```{r download-gse52576, eval = TRUE, cache = TRUE}
# Download GSE52576 from GEO
# Note: This is cached on first download to avoid repeated network calls
geo_gse <- "GSE52576"
message("Downloading ", geo_gse, " from GEO...")

tryCatch({
  gse <- getGEO(geo_gse, GSEMatrix = TRUE, AnnotGPL = TRUE, getGPL = FALSE)
  eset <- gse[[1]]  # Extract ExpressionSet object

  message("✓ Download successful")
  message("  Samples: ", ncol(eset))
  message("  Probes:  ", nrow(eset))
}, error = function(e) {
  stop("Failed to download ", geo_gse, ":\n", conditionMessage(e),
       "\nCheck internet connection and GEO availability.")
})
```

### Extract metadata and beta values

```{r extract-metadata}
# Extract phenotype data (metadata)
pdata <- pData(eset)
message("Available metadata columns:")
print(colnames(pdata))

# Display sample information
message("\nSample overview:")
head(pdata[, c("geo_accession", "title", "source_name_ch1")], n = 3)
```

### Format data for imprintomeR

```{r format-data}
# Extract beta values (methylation intensities, 0-1 scale)
beta_matrix <- exprs(eset)
message("Beta matrix dimensions: ", nrow(beta_matrix), " probes × ",
        ncol(beta_matrix), " samples")
message("Beta range: [", round(min(beta_matrix, na.rm = TRUE), 3), ", ",
        round(max(beta_matrix, na.rm = TRUE), 3), "]")

# Create metadata data.frame compatible with imprintomeR
meta <- data.frame(
  Sample_Name = colnames(beta_matrix),
  Sample_Group = pdata$source_name_ch1,  # Disease status or tissue type
  Disease_Status = pdata$characteristics_ch1.1,  # Detailed phenotype
  stringsAsFactors = FALSE
)

# Clean up Sample_Group (shorten for readability)
meta$Sample_Group <- gsub("^.*: ", "", meta$Sample_Group)

message("\nMetadata created:")
message("Samples: ", nrow(meta))
print(table(meta$Sample_Group))

message("\nFirst few samples:")
head(meta, n = 3)
```

### Validate data integrity

```{r validate-data}
# Check for missing values
pct_missing <- colSums(is.na(beta_matrix)) / nrow(beta_matrix) * 100
message("Missing values per sample:")
if (max(pct_missing) > 0) {
  print(head(sort(pct_missing, decreasing = TRUE)))
} else {
  message("  ✓ No missing values")
}

# Verify sample name alignment
if (all(colnames(beta_matrix) == meta$Sample_Name)) {
  message("✓ Sample names aligned between beta and metadata")
} else {
  stop("Sample name mismatch!")
}
```

---

## Phase 2: Load & Create ImprintomeSet

### Load data using imprintomeR utilities

```{r load-methqa}
# Use imprintomeR's LoadMetaBeta function to validate and intersect data
beta <- as.data.frame(beta_matrix)
beta <- cbind(NAME = rownames(beta), beta)
input <- LoadMetaBeta(meta, beta, probeset = NULL)

message("Loaded data:")
message("  Samples: ", ncol(input$beta))
message("  Probes:  ", nrow(input$beta))
message("  Metadata cols: ", paste(colnames(input$meta), collapse = ", "))

# Peek at data structure
message("\nBeta matrix (first 5 probes, 3 samples):")
print(round(input$beta[1:5, 1:3], 3))
```

### Create ImprintomeSet object

```{r create-imprintomeset}
# Load probeset annotations
probesets <- readRDS(
  system.file("extdata", "probesets_hg19.rds", package = "imprintomeR")
)

# Create analysis-ready ImprintomeSet
x <- ImprintomeSet(
  beta = input$beta,
  meta = input$meta,
  probeset = probesets[["selected"]],  # Use all probes; will be subset by runImprintome
  genome = "hg19",
  assay = "450K"  # GSE52576 uses Illumina 450K (close to EPIC v1)
)

# Verify object structure
message("ImprintomeSet created:")
message("  Samples: ", ncol(beta(x)))
message("  Probes: ", nrow(beta(x)))
message("  Genome: ", x@genome)
message("  Assay: ", x@assay)
print(summary(x))
```

---

## Phase 3: Run Workflow & Compute Statistics

### Execute imprintomeR analysis

```{r run-imprintome}
message("Running imprintome analysis...")
x <- runImprintome(x)
message("Analysis complete")

# Extract results
res <- results(x)
message("\nResults available:")
print(names(res))

# Display analysis summary
message("\nAnalysis summary:")
status <- res$AnalyzeImprintStatus.selected
print(head(status,n=5))

```

### Compute per-group statistics

```{r compute-stats}
# Summarize by Sample_Group
message("\nImprinting status by group:")
print(table(status[, c("Sample_Group", "Status")]))
```

### Review key metrics

```{r review-metrics}
# Distribution of key imprinting metrics
message("IDS (Imprinting Defect Score) distribution:")
print(summary(status$IDS))

message("\nStatus classification:")
print(table(status$Status))

message("\nAngle (parental origin deviation) distribution:")
print(summary(status$Angle))
```

---

## Phase 4: Visualizations - Complete Gallery

```{r run-visualizations, fig.width = 10, fig.height = 6}
sample_id_1 <- colnames(beta(x))[2]
message("Representative sample for single-sample plots: ", sample_id_1)

x <- runImprintomeVisualizations(
  x,
  probeset = "selected",
  sample_id = sample_id_1,
  prefix = "GSE52576",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)

attr(x, "visualization_manifest")
names(plots(x))
```

### Display selected stored plots

```{r display-stored-plots, fig.width = 10, fig.height = 6}
if (!is.null(plots(x)[["polar.selected"]])) {
  options(repr.plot.width = 8, repr.plot.height = 8)
  print(plots(x)[["polar.selected"]])
}

if (!is.null(plots(x)[["beeswarm_origin.selected"]])) {
   options(repr.plot.width = 12, repr.plot.height = 6)
  print(plots(x)[["beeswarm_origin.selected"]])
}

rainfall_name <- grep("^rainfall\\.", names(plots(x)), value = TRUE)[1]
if (!is.na(rainfall_name)) {
    options(repr.plot.width = 12, repr.plot.height = 6)
  print(plots(x)[[rainfall_name]])
}
```

### Visualizations by plot_type

The workflow helper can also run one plot backend at a time by setting `plot_types` explicitly.

#### Plot 1: Polar scatter

```{r plot-polar, fig.width = 8, fig.height = 8}
x <- runImprintomeVisualizations(
  x,
  plot_types = "polar",
  probeset = "selected",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "Sample_Group",
  prefix = "GSE52576_polar",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
options(repr.plot.width = 8, repr.plot.height = 8)
print(plots(x)[["polar.selected"]])
```

#### Plot 2: Beeswarm by origin

```{r plot-beeswarm-origin, fig.width = 12, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  plot_types = "beeswarm_origin",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  prefix = "GSE52576_beeswarm_origin",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
options(repr.plot.width = 12, repr.plot.height = 6)
print(plots(x)[["beeswarm_origin.selected"]])
```

#### Plot 3: Mirror density

```{r plot-mirror-density, fig.width = 10, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  plot_types = "mirror_density",
  probeset = "selected",
  prefix = "GSE52576_mirror_density",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
options(repr.plot.width = 10, repr.plot.height = 6)
print(plots(x)[["mirror_density.selected"]])
```

#### Plot 4: Circular heatmap

```{r plot-circular, fig.width = 10, fig.height = 10}
x <- runImprintomeVisualizations(
  x,
  plot_types = "circular_heatmap",
  probeset = "selected",
  sectionColumn = "Sample_Group",
  prefix = "GSE52576_circular_heatmap",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
options(repr.plot.width = 10, repr.plot.height = 10)
print(plots(x)[["circular.selected"]])
```

#### Plot 5: Beta heatmap by probe

```{r plot-heatmap-by-probe, fig.width = 10, fig.height = 8}
x <- runImprintomeVisualizations(
  x,
  plot_types = "heatmap_by_probe",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  annoColumn = "Sample_Group",
  prefix = "GSE52576_heatmap_by_probe",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
options(repr.plot.width = 10, repr.plot.height = 8)
print(plots(x)[["heatmap.selected"]])
```

#### Plot 6: Beta heatmap by gene

```{r plot-heatmap-by-gene, fig.width = 10, fig.height = 12}
x <- runImprintomeVisualizations(
  x,
  plot_types = "heatmap_by_gene",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  annoColumn = "Sample_Group",
  clusterRows = TRUE,
  clusterColumns = TRUE,
  prefix = "GSE52576_heatmap_by_gene",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
options(repr.plot.width = 10, repr.plot.height = 12)
print(plots(x)[["heatmap_by_gene.selected"]])
```

#### Plot 7: Radar plot

```{r plot-radar, fig.width = 10, fig.height = 10}
x <- runImprintomeVisualizations(
  x,
  plot_types = "radar",
  probeset = "selected",
  sample_id = sample_id_1,
  prefix = "GSE52576_radar",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
radar_name <- grep("^radar\\.", names(plots(x)), value = TRUE)[1]
if (!is.na(radar_name)) {
  options(repr.plot.width = 12, repr.plot.height = 12)
  print(plots(x)[[radar_name]])
}
```

#### Plot 8: Beeswarm by chromosome

```{r plot-beeswarm-chr, fig.width = 12, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  plot_types = "beeswarm_chr",
  probeset = "selected",
  sample_id = sample_id_1,
  prefix = "GSE52576_beeswarm_chr",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
beeswarm_chr_name <- grep("^beeswarm_chr\\.", names(plots(x)), value = TRUE)[1]
if (!is.na(beeswarm_chr_name)) {
  options(repr.plot.width = 12, repr.plot.height = 6)
  print(plots(x)[[beeswarm_chr_name]])
}
```

#### Plot 9: Rainfall plot

```{r plot-rainfall, fig.width = 12, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  plot_types = "rainfall",
  sample_id = sample_id_1,
  rainfall_probeset = "classifier3",
  prefix = "GSE52576_rainfall",
  store_plots = TRUE,
  save_plots = FALSE,
  verbose = TRUE
)
rainfall_name <- grep("^rainfall\\.", names(plots(x)), value = TRUE)[1]
if (!is.na(rainfall_name)) {
  options(repr.plot.width = 12, repr.plot.height = 6)
  print(plots(x)[[rainfall_name]])
}
```

---

## Phase 5: Results Interpretation & Case Study

### Summary statistics

```{r interpret-summary}
res <- results(x)
status <- res$AnalyzeImprintStatus.selected
normal_n <- sum(status$Status %in% "Normal", na.rm = TRUE)
total_n <- nrow(status)
normal_pct <- normal_n / total_n * 100
abnormal_pct <- 100 - normal_pct

message("=== OVERALL IMPRINTING SUMMARY ===")
message(sprintf("Normal imprinting:   %6.1f%% (%d/%d samples)",
                normal_pct, normal_n, total_n))
message(sprintf("Altered imprinting:  %6.1f%% (%d/%d samples)",
                abnormal_pct, total_n - normal_n, total_n))

message("\n=== BY SAMPLE GROUP ===")
if ("Sample_Group" %in% colnames(status)) {
  for (group in unique(status$Sample_Group)) {
    idx <- status$Sample_Group == group
    n <- sum(idx, na.rm = TRUE)
    normal <- sum(status$Status[idx] == "Normal", na.rm = TRUE)
    message(sprintf("%20s: %d samples, %d normal (%.1f%%)",
                    group, n, normal, 100 * normal / n))
  }
} else {
  message("Sample_Group column not available in status table.")
}
```

### Identify samples with altered imprinting

```{r identify-abnormal}
abnormal <- status %>% filter(!Status %in% "Normal")
sample_col <- intersect(c("Sample_Name", "Sample_ID"), colnames(status))[1]

if (nrow(abnormal) > 0) {
  message("=== ALTERED IMPRINTING SAMPLES ===")
  message("Sample count: ", nrow(abnormal))
  cols_to_show <- c(sample_col, intersect("Sample_Group", colnames(abnormal)),
                    "Status", "IDS", "Angle")
  print(abnormal %>% select(all_of(cols_to_show)) %>% head(n = 12))
} else {
  message("All samples show normal imprinting patterns by the selected IDS cutoff.")
}
```

### Phenotype-imprinting associations

```{r phenotype-associations}
if ("Sample_Group" %in% colnames(status) && length(unique(status$Sample_Group)) > 1) {
  message("Cross-tabulation: Sample_Group vs Imprinting Status")
  tbl <- table(status$Sample_Group, status$Status)
  print(tbl)

  group_summary <- status %>%
    group_by(Sample_Group, Status) %>%
    summarise(
      n = n(),
      median_IDS = median(IDS, na.rm = TRUE),
      median_Angle = median(Angle, na.rm = TRUE),
      .groups = "drop"
    )
  print(group_summary)
} else {
  message("Sample_Group is unavailable or has one level; skipping group cross-tabulation.")
}
```

### Key findings

```{r key-findings}
message("=== KEY BIOLOGICAL INSIGHTS ===")
message("1. Dataset: GSE52576, methylation study of imprinting disorders")
message("2. Platform: Illumina HumanMethylation450K")
message("3. Samples analyzed: ", total_n)
message("4. IDS range: [",
        round(min(status$IDS, na.rm = TRUE), 3), ", ",
        round(max(status$IDS, na.rm = TRUE), 3), "]")
message("5. Most common imprinting status: ",
        names(sort(table(status$Status), decreasing = TRUE))[1])
message("6. Compare Sample_Group summaries above to relate imprinting status to phenotype groups.")
```
---

## Phase 6: Export Results

```{r export-results, eval = FALSE}
manifest <- export(
  x,
  outdir = "GSE52576_imprintome_output",
  prefix = "GSE52576",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

head(manifest)
```

`export()` writes the full `ImprintomeSet` object as `<prefix>_imprintomeSet.rds`, plus `results(x)` and optionally `plots(x)`. Save extra summary tables only when you need custom reporting beyond the export manifest.

---

## Session Information & Reproducibility

```{r session-info}
# Reproducibility information
message("=== REPRODUCIBILITY ===")
message("R version: ", R.version$version.string)
message("imprintomeR version: ", packageVersion("imprintomeR"))
message("Execution date: ", Sys.Date())
message("Random seed: 2026")

sessionInfo()
```

---

## Jupyter Notebook Conversion Guide

This vignette is designed for conversion to Jupyter notebooks. To convert:

```bash
# 1. Render this Rmd to HTML (in R):
rmarkdown::render('vignettes/GSE52576.Rmd')

# 2. Convert HTML to Jupyter notebook (requires nbconvert):
jupyter nbconvert --to notebook GSE52576.html --output GSE52576.ipynb

# 3. Or use Jupytext to convert Rmd directly to .ipynb:
jupytext --to notebook vignettes/GSE52576.Rmd
```

**Tips for Jupyter users:**
- Each code block above is a separate cell (copy/paste ready)
- Markdown sections become markdown cells
- Uncomment `outFile` parameters in plot calls to save PDFs
- Use `cache=TRUE` on the download cell to avoid re-downloading
- All plots can be viewed inline; no file system navigation required

---

## References

- **GSE52576**: [NCBI GEO record](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52576)
- **imprintomeR documentation**: See package vignettes
- **GEOquery**: Bioconductor package for GEO data access
- **Imprinting Control Regions**: See `?setup_imprintome_data` for probeset annotations

---

## License & Attribution

This vignette uses publicly available data from NCBI Gene Expression Omnibus (GEO).
Please cite the original GSE52576 study and imprintomeR when using this analysis.

