---
title: "Real-World Imprinting Analysis: GSE52576 Dataset"
subtitle: "Download, preprocess, and analyze methylation data from a public GEO dataset"
author: "imprintomeR contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 10
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{GSE52576 Real-World Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "GSE52576_",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This vignette demonstrates imprintomeR on **GSE52576**, a real-world methylation dataset from the Gene Expression Omnibus (GEO). You will:

1. **Download** normalized beta values from GEO using GEOquery
2. **Create** an ImprintomeSet analysis-ready object
3. **Analyze** imprinting status across samples
4. **Visualize** results with all available plot types
5. **Interpret** biological patterns and aberrant imprinting

The workflow is optimized for **reproducibility** and **Jupyter notebook conversion**.

---

## Phase 1: Setup & Data Download

### Load libraries

```{r load-libraries, message = FALSE, warning = FALSE}
library(imprintomeR)
library(GEOquery)
library(tidyverse)
library(ComplexHeatmap)
library(ggplot2)
library(knitr)

# For reproducibility
set.seed(2026)
```

### Download GSE52576 from GEO

GSE52576 is a publicly available dataset studying Beckwith-Wiedemann syndrome (BWS) and imprinting defects using Illumina 450K methylation arrays.

```{r download-gse52576, eval = TRUE, cache = TRUE}
# Download GSE52576 from GEO
# Note: This is cached on first download to avoid repeated network calls
geo_gse <- "GSE52576"
message("Downloading ", geo_gse, " from GEO...")

tryCatch({
  gse <- getGEO(geo_gse, GSEMatrix = TRUE, AnnotGPL = TRUE, getGPL = FALSE)
  eset <- gse[[1]]  # Extract ExpressionSet object
  
  message("✓ Download successful")
  message("  Samples: ", ncol(eset))
  message("  Probes:  ", nrow(eset))
}, error = function(e) {
  stop("Failed to download ", geo_gse, ":\n", conditionMessage(e), 
       "\nCheck internet connection and GEO availability.")
})
```

### Extract metadata and beta values

```{r extract-metadata}
# Extract phenotype data (metadata)
pdata <- pData(eset)
message("Available metadata columns:")
print(colnames(pdata))

# Display sample information
message("\nSample overview:")
head(pdata[, c("geo_accession", "title", "source_name_ch1")], n = 3)
```

### Format data for imprintomeR

```{r format-data}
# Extract beta values (methylation intensities, 0-1 scale)
beta_matrix <- exprs(eset)
message("Beta matrix dimensions: ", nrow(beta_matrix), " probes × ", 
        ncol(beta_matrix), " samples")
message("Beta range: [", round(min(beta_matrix, na.rm = TRUE), 3), ", ",
        round(max(beta_matrix, na.rm = TRUE), 3), "]")

# Create metadata data.frame compatible with imprintomeR
meta <- data.frame(
  Sample_Name = colnames(beta_matrix),
  Sample_Group = pdata$source_name_ch1,  # Disease status or tissue type
  Disease_Status = pdata$characteristics_ch1.1,  # Detailed phenotype
  stringsAsFactors = FALSE
)

# Clean up Sample_Group (shorten for readability)
meta$Sample_Group <- gsub("^.*: ", "", meta$Sample_Group)

message("\nMetadata created:")
message("Samples: ", nrow(meta))
print(table(meta$Sample_Group))

message("\nFirst few samples:")
head(meta, n = 3)
```

### Validate data integrity

```{r validate-data}
# Check for missing values
pct_missing <- colSums(is.na(beta_matrix)) / nrow(beta_matrix) * 100
message("Missing values per sample:")
if (max(pct_missing) > 0) {
  print(head(sort(pct_missing, decreasing = TRUE)))
} else {
  message("  ✓ No missing values")
}

# Verify sample name alignment
if (all(colnames(beta_matrix) == meta$Sample_Name)) {
  message("✓ Sample names aligned between beta and metadata")
} else {
  stop("Sample name mismatch!")
}
```

---

## Phase 2: Load & Create ImprintomeSet

### Load data using imprintomeR utilities

```{r load-methqa}
# Use imprintomeR's LoadMetaBeta function to validate and intersect data
input <- LoadMetaBeta(meta, beta_matrix, probeset = NULL)

message("Loaded data:")
message("  Samples: ", ncol(input$beta))
message("  Probes:  ", nrow(input$beta))
message("  Metadata cols: ", paste(colnames(input$meta), collapse = ", "))

# Peek at data structure
message("\nBeta matrix (first 5 probes, 3 samples):")
print(round(input$beta[1:5, 1:3], 3))
```

### Create ImprintomeSet object

```{r create-imprintomeset}
# Create analysis-ready ImprintomeSet
x <- as.ImprintomeSet(
  beta = input$beta,
  meta = input$meta,
  probeset = NULL,  # Use all probes; will be subset by runImprintome
  genome = "hg19",
  assay = "EPIC"  # GSE52576 uses Illumina 450K (close to EPIC v1)
)

# Verify object structure
message("ImprintomeSet created:")
message("  Samples: ", ncol(beta(x)))
message("  Probes: ", nrow(beta(x)))
message("  Genome: ", x@genome)
message("  Assay: ", x@assay)
print(summary(x))
```

---

## Phase 3: Run Workflow & Compute Statistics

### Execute imprintomeR analysis

```{r run-imprintome}
message("Running imprintome analysis...")
x <- runImprintome(x)
message("✓ Analysis complete")

# Extract results
res <- results(x)
message("\nResults available:")
print(names(res))

# Display analysis summary
message("\nAnalysis summary:")
print(head(res$status, n = 5))
```

### Compute per-group statistics

```{r compute-stats}
# Summarize by Sample_Group
stat_summary <- res$status %>%
  group_by(Sample_Group = meta(x)$Sample_Group[match(Sample_ID, Sample_Name)]) %>%
  summarise(
    n_samples = n(),
    n_normal = sum(Mechanism == "NORMAL"),
    n_disomy = sum(Mechanism %in% c("DISOMY", "DISOMY-like")),
    n_upd = sum(grepl("UPD|matUPD|patUPD", Mechanism)),
    pct_abnormal = round(100 * (1 - sum(Mechanism == "NORMAL") / n()), 1),
    .groups = "drop"
  )

message("\nImprinting status by group:")
print(stat_summary)
```

### Review key metrics

```{r review-metrics}
# Distribution of key imprinting metrics
message("IDS (Imprinting Defect Score) distribution:")
print(summary(res$status$IDS))

message("\nMechanism classification:")
print(table(res$status$Mechanism))

message("\nAngle (parental origin deviation) distribution:")
print(summary(res$status$Angle))
```

---

## Phase 4: Visualizations — Complete Gallery

### Plot 1: Polar scatter (IDS vs Angle)

The **polar plot** is the primary visualization—it shows each sample's imprinting deviation in 2D space, colored by mechanism classification.

```{r plot-polar, fig.width = 8, fig.height = 8}
p_polar <- plot(
  x,
  plot_type = "polar",
  outFile = NULL  # Display only; set to filename to save
)
print(p_polar)
message("✓ Polar plot complete")
```

### Plot 2: Beta heatmap (methylation by probeset)

The **heatmap** shows normalized methylation levels (beta values) across all probes and samples, organized by imprinting mechanism.

```{r plot-heatmap, fig.width = 10, fig.height = 8}
p_heatmap <- plot(
  x,
  plot_type = "heatmap",
  outFile = NULL
)
message("✓ Heatmap complete")
```

### Plot 3: Beeswarm by origin (cohort-level methylation)

The **beeswarm plot** shows methylation distribution for each sample, faceted by parental origin (paternal vs maternal imprinting).

```{r plot-beeswarm-origin, fig.width = 12, fig.height = 6}
p_beeswarm_origin <- plot(
  x,
  plot_type = "beeswarm_origin",
  outFile = NULL
)
message("✓ Beeswarm by origin plot complete")
```

### Plot 4: Circular heatmap (complex multi-dimensional view)

The **circular heatmap** integrates probeset annotations (chromosome, imprinting type) with methylation and statistical significance.

```{r plot-circular, fig.width = 10, fig.height = 10}
p_circular <- plot(
  x,
  plot_type = "circular_heatmap",
  outFile = NULL
)
message("✓ Circular heatmap complete")
```

### Plot 5: Rainfall plot (sample-level detail)

The **rainfall plot** shows methylation deviations (IDI) across genomic positions for a single representative sample. Select sample 2 for demonstration:

```{r plot-rainfall, fig.width = 12, fig.height = 6}
sample_id_1 <- colnames(beta(x))[2]  # Select second sample
message("Generating rainfall plot for sample: ", sample_id_1)

p_rainfall <- plot(
  x,
  plot_type = "rainfall",
  sample_id = sample_id_1,
  probeset = "classifier3",
  outFile = NULL
)
message("✓ Rainfall plot complete")
```

### Plot 6: Radar plot (multi-dimensional sample summary)

The **radar plot** displays a single sample's imprinting profile across multiple dimensions (IDS, Angle, IDI summary).

```{r plot-radar, fig.width = 8, fig.height = 8}
p_radar <- plot(
  x,
  plot_type = "radar",
  sample_id = sample_id_1,
  probeset = "selected",
  outFile = NULL
)
message("✓ Radar plot complete")
```

### Plot 7: Beeswarm by chromosome (spatially organized)

The **chromosome beeswarm plot** shows methylation distribution for a single sample, faceted by chromosome location of imprinted sites.

```{r plot-beeswarm-chr, fig.width = 12, fig.height = 6}
p_beeswarm_chr <- plot(
  x,
  plot_type = "beeswarm_chr",
  sample_id = sample_id_1,
  outFile = NULL
)
message("✓ Beeswarm by chromosome plot complete")
```

---

## Phase 5: Results Interpretation & Case Study

### Summary statistics

```{r interpret-summary}
# Overall imprinting status
normal_pct <- sum(res$status$Mechanism == "NORMAL") / nrow(res$status) * 100
abnormal_pct <- 100 - normal_pct

message("=== OVERALL IMPRINTING SUMMARY ===")
message(sprintf("Normal imprinting:    %6.1f%% (%d/%d samples)", 
                normal_pct, sum(res$status$Mechanism == "NORMAL"), nrow(res$status)))
message(sprintf("Abnormal imprinting:  %6.1f%% (%d/%d samples)", 
                abnormal_pct, nrow(res$status) - sum(res$status$Mechanism == "NORMAL"), 
                nrow(res$status)))

# By sample group
message("\n=== BY SAMPLE GROUP ===")
for (group in unique(meta(x)$Sample_Group)) {
  idx <- meta(x)$Sample_Group == group
  n <- sum(idx)
  normal <- sum(res$status$Mechanism[idx] == "NORMAL")
  message(sprintf("%20s: %d samples, %d normal (%.1f%%)", 
                  group, n, normal, 100*normal/n))
}
```

### Identify samples with aberrant imprinting

```{r identify-abnormal}
abnormal <- res$status %>% filter(Mechanism != "NORMAL")

if (nrow(abnormal) > 0) {
  message("=== ABERRANT IMPRINTING SAMPLES ===")
  message("Sample count: ", nrow(abnormal))
  print(abnormal %>% select(Sample_ID, Mechanism, IDS, Angle) %>% head(n = 10))
} else {
  message("✓ All samples show normal imprinting patterns")
}
```

### Phenotype-imprinting associations

```{r phenotype-associations}
# Add phenotype info to results for cross-reference
res_annotated <- res$status %>%
  mutate(
    Sample_Name = Sample_ID,
    Sample_Group = meta(x)$Sample_Group[match(Sample_ID, meta(x)$Sample_Name)]
  )

# Compare disease status vs imprinting
if (length(unique(res_annotated$Sample_Group)) > 1) {
  message("Cross-tabulation: Sample_Group vs Imprinting Mechanism")
  tbl <- table(res_annotated$Sample_Group, res_annotated$Mechanism)
  print(tbl)
}
```

### Key findings

```{r key-findings}
message("=== KEY BIOLOGICAL INSIGHTS ===")
message("1. Dataset: GSE52576 (methylation study of imprinting disorders)")
message("2. Platform: Illumina 450K methylation array")
message("3. Samples analyzed: ", nrow(res$status))
message("4. Geographic distribution: IDS range [", 
        round(min(res$status$IDS), 3), ", ",
        round(max(res$status$IDS), 3), "]")
message("5. Most common mechanism: ", 
        names(sort(table(res$status$Mechanism), decreasing = TRUE))[1])
message("\n→ See plots above for detailed visualization of these patterns")
```

---

## Phase 6: Export & Save Results

### Export to Excel workbooks

```{r export-xlsx}
# Define output directory
outdir <- "GSE52576_output"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

# Export ImprintomeSet results
export(
  x,
  outdir = outdir,
  format = c("xlsx", "rds"),
  prefix = "GSE52576"
)

message("✓ Results exported to: ", outdir)
message("  - GSE52576_*_main.xlsx (primary results)")
message("  - GSE52576_*_extra.xlsx (supplementary data)")
message("  - GSE52576_imprintomeset.rds (full object)")
```

### Export results to TSV

```{r export-txt}
# Write results table
results_file <- file.path(outdir, "GSE52576_imprintome_results.txt")
write.table(
  res$status,
  file = results_file,
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE
)

message("✓ Results table: ", results_file)

# Write summary statistics
summary_file <- file.path(outdir, "GSE52576_summary_statistics.txt")
write.table(
  stat_summary,
  file = summary_file,
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE
)

message("✓ Summary statistics: ", summary_file)
```

### List all output files

```{r list-outputs}
output_files <- list.files(outdir, full.names = FALSE)
message("=== OUTPUT FILES ===")
message("Directory: ", outdir)
for (f in sort(output_files)) {
  fsize <- file.size(file.path(outdir, f))
  message(sprintf("  %-40s %8.1f KB", f, fsize / 1024))
}
```

---

## Session Information & Reproducibility

```{r session-info}
# Reproducibility information
message("=== REPRODUCIBILITY ===")
message("R version: ", R.version$version.string)
message("imprintomeR version: ", packageVersion("imprintomeR"))
message("Execution date: ", Sys.Date())
message("Random seed: 2026")

sessionInfo()
```

---

## Jupyter Notebook Conversion Guide

This vignette is designed for conversion to Jupyter notebooks. To convert:

```bash
# 1. Render this Rmd to HTML (in R):
rmarkdown::render('vignettes/GSE52576.Rmd')

# 2. Convert HTML to Jupyter notebook (requires nbconvert):
jupyter nbconvert --to notebook GSE52576.html --output GSE52576.ipynb

# 3. Or use Jupytext to convert Rmd directly to .ipynb:
jupytext --to notebook vignettes/GSE52576.Rmd
```

**Tips for Jupyter users:**
- Each code block above is a separate cell (copy/paste ready)
- Markdown sections become markdown cells
- Uncomment `outFile` parameters in plot calls to save PDFs
- Use `cache=TRUE` on the download cell to avoid re-downloading
- All plots can be viewed inline; no file system navigation required

---

## References

- **GSE52576**: [NCBI GEO record](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52576)
- **imprintomeR documentation**: See package vignettes
- **GEOquery**: Bioconductor package for GEO data access
- **Imprinting Control Regions**: See `?setup_imprintome_data` for probeset annotations

---

## License & Attribution

This vignette uses publicly available data from NCBI Gene Expression Omnibus (GEO). 
Please cite the original GSE52576 study and imprintomeR when using this analysis.

