---
title: "ImprintomeSet Visualization Gallery"
author: "imprintomeR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ImprintomeSet Visualization Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates the full visualization and export workflow after completing
imprinting analysis with `runImprintome()`.

Visualization in imprintomeR follows the preprocessing-first architecture:

1. **MethQcSet**: QC preprocessing (see *ImprintomeSet Workflow Quickstart*)
2. **ImprintomeSet**: Analysis-ready container for imprinting studies (this vignette)
3. **Results & Plots**: Stored in `results(x)` and `plots(x)` slots

The examples below are organized by visualization level:

- sample-level (polar, radar, rainfall)
- probe-level (beeswarm, violin, mirror density)
- cohort-level (heatmap, cor_heatmap, circular_heatmap)
- chromosome-level (beeswarm_chr)

For object construction and QC preprocessing, see
*ImprintomeSet Workflow Quickstart*.

## 1. Start from a QC-Approved ImprintomeSet Object

Your `ImprintomeSet` object `x` should come from **either**:

**Option A: QC-preprocessed data via MethQcSet**

```{r, eval = FALSE}
# Load metadata: tab-delimited with at least Sample_Name and Sample_Group.
# For QC workflow (raw IDAT files): also include Basename column (IDAT path prefixes).
# For direct analysis (QC already done): Basename is not needed.
meta <- LoadMeta("meta.tsv")

# Option A1: platform unknown — detect first
meta <- check_platform(meta)
meta_epic <- meta[meta$Platform == "EPIC", ]
qcset <- runMethQC(meta_epic)  # platform auto-resolved from meta$Platform

# Option A2: platform known — pass directly
# qcset <- runMethQC(meta, platform = "EPIC")

export(qcset, outdir = "qc_output", format = c("xlsx", "rds"))
# xlsx: two workbooks for organized reporting:
#   1. qc_table_main.xlsx — metadata (+ Platform, Final.QC), QC_matrix, Ctrl_matrix, statistics
#   2. qc_table_extra.xlsx — QC_metrics (thresholds), contamination, recall_rate, predUniqDonor_ID
# rds: Full MethQcSet object for programmatic access

# QC visualizations
p_bar  <- plot(qcset, type = "qc_bar")                          # PASS/FAIL bar
p_int  <- plot(qcset, type = "intensity",  icutoff = 11)        # intensity scatter
p_dp   <- plot(qcset, type = "detection_pval", pcutoff = 0.05)  # avg det p-val
p_cov  <- plot(qcset, type = "probe_coverage")                  # % probes detected
p_sex  <- plot(qcset, type = "predicted_sex")                   # predicted sex
p_ctrl <- plot(qcset, type = "ctrl_metrics")                    # ewastools (if present)
plot(qcset, type = "ctrl_metrics_detail",                       # per-metric detail PDFs
     outFile = "qc_ctrl_matrix_detail")

# EPICv2 only: aggregate replicate probes before conversion (required step)
# as.ImprintomeSet() will error if this is skipped for EPICv2 data.
if (qcset@platform == "EPICv2") {
  qcset <- aggregate_probes(qcset)
}

# Convert the QC-clean MethQcSet to ImprintomeSet
probesets <- readRDS(system.file("extdata", "probesets_hg19.rds", package = "imprintomeR"))
x <- as.ImprintomeSet(
  qcset,
  probeset = probesets[["selected"]],
  genome = "hg19"
)
```

**Option B: Pre-cleaned data loaded directly**

```{r, eval = FALSE}
# If you already have QC-cleaned data, skip MethQcSet
# Note: Basename column is not needed since QC step is skipped.
meta <- LoadMeta("meta.tsv")  # Only needs Sample_Name and Sample_Group
beta <- LoadBeta("cleaned_beta.tsv")

x <- ImprintomeSet(
  beta = beta,
  meta = meta,
  probeset = probesets[["selected"]],
  genome = "hg19",
  assay = "EPICv1"
)
```

Then run analysis on your analysis-ready `ImprintomeSet`:

```{r, eval = FALSE}
library(imprintomeR)

# x is a valid ImprintomeSet object (QC-approved data)

# Run analysis on one or more probesets
x <- runImprintome(x, probeset = "selected", ids_cutoff = 0.2)
x <- runImprintome(x, probeset = "classifier3", ids_cutoff = 0.2,
                   result_name = "AnalyzeImprintStatus.classifier3")

names(results(x))
```

Extract slots once for plotting convenience:

```{r, eval = FALSE}
beta_x <- beta(x)
meta_x <- meta(x)
res_x <- results(x)[["AnalyzeImprintStatus.selected"]]
```

## 2. Generic S4 Plot Method

The S4 `plot()` method supports explicit backend dispatch via `plot_type`.

Available `plot_type` values are:

- `auto`
- `polar`
- `mirror_density`
- `beeswarm`
- `beeswarm_origin`
- `beeswarm_chr`
- `violin`
- `heatmap`
- `circular_heatmap`
- `cor_heatmap`
- `rainfall`
- `radar`

### 2.1 Auto Dispatch

`auto` first returns the latest stored plot from `plots(x)` when present;
otherwise, it falls back to polar plotting from IDS/Angle results.

```{r, eval = FALSE}
p_auto <- plot(
  x,
  plot_type = "auto",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "Sample_Group",
  outFile = "auto_dispatch.pdf"
)
```

### 2.2 Explicit Dispatch

```{r, eval = FALSE}
p_polar_s4 <- plot(
  x,
  plot_type = "polar",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "Sample_Group",
  title = "Selected Probeset Polar",
  palette = "default",
  alpha = 0.5,
  outFile = "selected_polar.pdf"
)
```

### 2.3 Compact All-Plot-Type Demo

```{r, eval = FALSE}
sample_id1 <- colnames(beta(x))[2]

p_auto <- plot(x, plot_type = "auto", result_name = "AnalyzeImprintStatus.selected",
               colorColumn = "Sample_Group", outFile = "demo_auto.pdf")
p_polar <- plot(x, plot_type = "polar", result_name = "AnalyzeImprintStatus.selected",
                colorColumn = "Sample_Group", outFile = "demo_polar.pdf")
p_mirror <- plot(x, plot_type = "mirror_density", probeset = "selected",
                 SAMPLEID = "Sample_Name", outFile = "demo_mirror_density.pdf")
p_bee <- plot(x, plot_type = "beeswarm", SAMPLEID = "Sample_Name",
        probeset = "selected",
              outFile = "demo_beeswarm.pdf")
p_bee_origin <- plot(x, plot_type = "beeswarm_origin", SAMPLEID = "Sample_Name",
         probeset = "selected", title = "Beeswarm Origin (Cohort)",
         outFile = "demo_beeswarm_origin.pdf")
p_bee_chr <- plot(x, plot_type = "beeswarm_chr", probeset = "selected",
      sample_id = sample_id1, chr = "chr11",
      title = "Beeswarm by Chromosome", outFile = "demo_beeswarm_chr.pdf")
p_vln <- plot(x, plot_type = "violin", SAMPLEID = "Sample_Name",
        probeset = "selected",
              outFile = "demo_violin.pdf")
p_heatmap_by_probe <- plot(x, plot_type = "heatmap_by_probe", SAMPLEID = "Sample_Name",
          probeset = "selected",
                  outFile = "demo_heatmap_by_probe.pdf")

p_heatmap_by_gene <- plot(  x,  plot_type = "heatmap_by_gene",  probeset = "selected",
        SAMPLEID = "Sample_Name",  annoColumn = "Sample_Group",
        clusterRows = TRUE,  clusterColumns = TRUE,
        outFile = "demo_heatmap_by_gene.pdf")

p_circular <- plot(x, plot_type = "circular_heatmap", SAMPLEID = "Sample_Name",
                   sectionColumn = "Sample_Group", probeset = "selected",
                   outFile = "demo_circular_heatmap.pdf")
cor_mat <- plot(x, plot_type = "cor_heatmap", SAMPLEID = "Sample_Name",
                probeset = "selected",
                prefix = "demo")
outFile <- paste0("imprintomeset_quickstart_rainfall_", sample_id1, ".pdf")
p_rain <- plot(x, plot_type = "rainfall", sample_id = sample_id1,
               probeset = "classifier3", outFile = outFile)
outFile <- paste0("imprintomeset_quickstart_radar_", sample_id1, ".pdf")
p_radar <- plot(x, plot_type = "radar", probeset = "selected", sample_id = sample_id1,
                title = "ImprintomeSet Radar", outFile = outFile)
```

## 3. Sample-Level Visualizations

### 3.1 Polar Plot

```{r, eval = FALSE}
p_polar <- plot(
  x,
  plot_type = "polar",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "Sample_Group",
  title = "ImprintomeSet Polar",
  palette = "default",
  alpha = 0.5,
  outFile = "polar_direct.pdf"
)
```

### 3.2 Mirror Density

```{r, eval = FALSE}
p_mirror <- plot(
  x,
  plot_type = "mirror_density",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  outFile = "mirror_density.pdf"
)
```

### 3.3 Radar Plot

```{r, eval = FALSE}
sample_id1 <- colnames(beta(x))[2]
outFile <- paste0("imprintomeset_quickstart_radar_", sample_id1, ".pdf")
p_radar <- plot(
  x,
  plot_type = "radar",
  probeset = "selected",
  sample_id = sample_id1,
  title = "ImprintomeSet Radar",
  outFile = outFile
)
```

## 4. Probe-Level Visualizations

### 4.1 Violin Plot

```{r, eval = FALSE}
p_vln <- plot(
  x,
  plot_type = "violin",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  outFile = "beta_violin.pdf"
)
```

### 4.2 Beeswarm Plot

```{r, eval = FALSE}
p_bee <- plot(
  x,
  plot_type = "beeswarm",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  outFile = "beta_beeswarm.pdf"
)
```

### 4.3 Single-Chromosome Beeswarm

```{r, eval = FALSE}
p_bee_chr11 <- imprintomeR:::BetaBeePlot_single_chr(
  beta = beta_x,
  meta = meta_x,
  SAMPLEID = "Sample_Name",
  chr = "chr11",
  probeset = "chr11p15",
  outFile = "beeswarm_chr11.pdf"
)
```

### 4.4 Origin-Split Beeswarm

Use the ImprintomeSet dispatcher (`plot_type = "beeswarm_origin"`) for cohort-wide
origin split plots.

```{r, eval = FALSE}
p_bee_origin <- plot(
  x,
  plot_type = "beeswarm_origin",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  title = "Beeswarm Origin (Cohort)",
  outFile = "beeswarm_origin_dispatch.pdf"
)
```

### 4.5 Single-Sample Beeswarm Faceted by Chromosome

```{r, eval = FALSE}
sample_id1 <- colnames(beta(x))[2]
p_bee_chr <- plot(
  x,
  plot_type = "beeswarm_chr",
  probeset = "selected",
  sample_id = sample_id1,
  chr = "chr11",
  title = "Beeswarm by Chromosome (Origin Split)",
  outFile = "beeswarm_chr_dispatch.pdf"
)
```

The subtitle reports the selected sample, probeset, and chromosome scope.
Within each chromosome panel, maternal/paternal points are separated and median
summary markers are overlaid.

## 5. Cohort-Level Visualizations

### 5.1 Correlation Heatmap

```{r, eval = FALSE}
cor_mat <- plot(
  x,
  plot_type = "cor_heatmap",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  prefix = "cohort"
)
```

### 5.2 Heatmap_by_probe (Probe-Level Heatmap)

Displays individual Illumina probe IDs (e.g., cg00000029) as rows and samples as columns with their methylation values color-coded.

```{r, eval = FALSE}
plot(
  x,
  plot_type = "heatmap_by_probe",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  outFile = "beta_heatmap.pdf"
)
```

### 5.3 Circular Heatmap

Circular heatmap displays methylation values arranged in a circular layout. Each chromosome forms a radial sector, 
with probes arranged clockwise. Sample groups are rendered as separate "pie slices" around the center white circle,
allowing visual comparison of methylation patterns across grouped samples and chromosomes simultaneously.

**Parameters:**

- `sectionColumn`: metadata column for grouping samples (default `"Sample_Group"`). If the column is missing, 
  samples are grouped by `Sample_Name` (one section per sample). If both are missing, all samples appear in one section.
- `probeset`: probeset name (default `"selected"`)
- `SAMPLEID`: metadata column for sample labeling

**Features:**

- Shows probes × samples with chromosome organization
- Color scale: blue (unmethylated, β ≈ 0) → white (hemimethylated, β ≈ 0.5) → red (methylated, β ≈ 1)
- Automatic adaptive sizing: larger canvas (12×10 inches) accommodates 4+ sample groups
- Section labels positioned in the inner white circle for clarity

```{r, eval = FALSE}
# Group by sample group
plot(
  x,
  plot_type = "circular_heatmap",
  SAMPLEID = "Sample_Name",
  sectionColumn = "Sample_Group",
  probeset = "selected",
  outFile = "beta_circular_heatmap.pdf"
)
```



### 5.4 Heatmap Aggregated by Gene Symbol

This visualization aggregates probe-level beta values to gene-level summaries 
by grouping probes with the same `Closest_TSS_gene_name` from the selected probeset. 
Within each gene, probes are aggregated per sample using the **median** function.

**Features:**

- Shows gene symbols as rows and samples as columns
- Robust aggregation: reduces ~865k probes (EPICv2) → ~50-200 unique genes per probeset
- Clustered by default: genes clustered via Euclidean distance (Ward linkage)
- Top annotation: sample groups with automatic color assignment
- Color scale: blue (unmethylated) → white (hemimethylated) → red (methylated)

**Parameters:**

- `probeset`: probeset name (default `"selected"`)
- `SAMPLEID`: metadata column for sample labels
- `annoColumn`: annotation column for top heatmap header (default `"Sample_Group"`)
- `clusterRows`: hierarchical cluster genes? (default `TRUE`)
- `clusterColumns`: hierarchical cluster samples? (default `TRUE`)
- `outFile`: output PDF file path

```{r, eval = FALSE}
plot(
  x,
  plot_type = "heatmap_by_gene",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  annoColumn = "Sample_Group",
  clusterRows = TRUE,
  clusterColumns = TRUE,
  outFile = "beta_heatmap_by_gene.pdf"
)
```


## 6. Chromosome and Genome Visualizations

### 6.1 Chromosome Beeswarm Panels

```{r, eval = FALSE}
p_bee_chr_all <- plot(
  x,
  plot_type = "beeswarm_chr",
  probeset = "selected",
  sample_id = sample_id1,
  chr = NULL,
  title = "Beeswarm by Chromosome (All Chromosomes)",
  outFile = "beeswarm_by_chromosome.pdf"
)
```

### 6.2 Rainfall Plot

```{r, eval = FALSE}
sample_id1 <- colnames(beta(x))[2]
outFile <- paste0("imprintomeset_quickstart_rainfall_", sample_id1, ".pdf")

p_rain <- plot(
  x,
  plot_type = "rainfall",
  sample_id = sample_id1,
  probeset = "classifier3",
  outFile = outFile
)
```

### 6.3 Chromosome vs Others Beeswarm

```{r, eval = FALSE}
probesets <- readRDS(system.file("extdata", "probesets_hg19.rds", package = "imprintomeR"))
ps <- probesets[["classifier2"]]
common_probes <- intersect(ps$NAME, rownames(beta_x))

dat_chr <- beta_x[common_probes, , drop = FALSE]
dat_chr$chr <- ifelse(ps[common_probes, "CHR"] %in% c("chr11", "11"), "chr11", "others")
dat_chr$CATEGORY <- ps[common_probes, "ORIGIN"]

imprintomeR:::Beeplot_chr_vs_other(
  dat = dat_chr,
  meta = meta_x,
  group.by = "chr",
  color.by = "CATEGORY",
  outFile = "beeplot_chr11_vs_others.pdf"
)
```

### 6.4 Batch Chromosome vs Others Convenience Wrapper

```{r, eval = FALSE}
imprintomeR:::Beeplot_chr_vs_other_single(
  input = list(beta = beta_x, meta = meta_x),
  chrs = "chr1,chr11,chr20",
  prefix = "chr_vs_other_batch",
  probeset = "classifier2"
)
```

## 7. Persist Plot Objects Inside ImprintomeSet

```{r, eval = FALSE}
p_list <- plots(x)
p_list[["polar.selected"]] <- p_polar
p_list[["mirror.selected"]] <- p_mirror
p_list[["radar.sample1"]] <- p_radar
p_list[["beeswarm.cohort"]] <- p_bee
p_list[["beeswarm.origin.cohort"]] <- p_bee_origin
p_list[["beeswarm.chr.sample1"]] <- p_bee_chr
p_list[["violin.cohort"]] <- p_vln
p_list[["rainfall.sample1"]] <- p_rain
plots(x) <- p_list

names(plots(x))

# Access stored plot objects
p_origin_from_object <- plots(x)[["beeswarm.origin.cohort"]]
p_chr_from_object <- plots(x)[["beeswarm.chr.sample1"]]

p_origin_from_object
p_chr_from_object
```

## 8. Summarize and Export

```{r, eval = FALSE}
s <- summarize(x)
s$object
s$results
s$qc
s$plots

manifest <- export(
  x,
  outdir = "imprintome_export",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

head(manifest)
```

`export()` writes result tables and stored plot objects. QC tables remain available through `qc(x)` and `summarize(x)$qc`.

**For MethQcSet export**, the output includes a `summary.txt` file with an automatically-generated "Notes" section:

```{r, eval = FALSE}
export(qcset, outdir = "qc_output", format = c("xlsx", "rds"))

# Inspect the summary.txt file to see which sheets were created:
readLines("qc_output/methqcset_summary.txt")
```

The Notes section documents any optional sheets that were not created due to missing underlying data (e.g., missing `ctrl_metrics`, `contamination`, etc.). This provides transparency about data availability during export.

## 9. Notes

- Several plotting functions write side-effect files and also return plot objects.
- For large cohorts, disable per-sample side folders where optional (`pdfFolder = NULL`).
- Keep `Sample_Name` alignment between `beta(x)` and `meta(x)`.
- Advanced, non-dispatch plots can still be called directly when needed.

## Session Info

```{r, eval = FALSE}
sessionInfo()
```
