---
title: "imprintomeR Workflow"
author: "imprintomeR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{imprintomeR Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`imprintomeR` provides analysis and visualization utilities for genomic imprinting
from methylation array data. This vignette shows the complete preprocessing-first workflow:

1. **MethQcSet**: QC preprocessing for single-platform raw data
2. **Function-based analysis**: `AnalyzeImprintStatus()` and related functions
3. **Object-oriented analysis**: `ImprintomeSet` container and `runImprintome()`

Core metrics follow the package scientific contract:

- IDS: distance from the balanced state `(0.5, 0.5)`
- Angle: direction from paternal/maternal deviations relative to `0.5`
- IDI: `(beta - 0.5) \times 2`

For dedicated object-first guides, see:

- *ImprintomeSet Workflow Quickstart*
- *ImprintomeSet Visualization Gallery*

## 1. QC Preprocessing (Optional if Data Already Cleaned)

**If you have RAW data:**

Load metadata and detect platform from IDAT files:

```{r, eval = FALSE}
library(imprintomeR)

# meta must have Basename column (IDAT path prefixes)
meta <- LoadMeta("meta.tsv")
```

`runMethQC()` reads IDAT files directly — no separate beta loading step needed.

If the platform is **unknown or mixed**, detect it first:

```{r, eval = FALSE}
# Reads one IDAT per sample to infer platform from probe count
# Appends Platform, Array_Size, Manifest, Status columns to meta
meta <- check_platform(meta)
table(meta$Platform)  # inspect; subset before calling runMethQC

meta_epic <- meta[meta$Platform == "EPIC", ]
qcset <- runMethQC(meta_epic)  # platform auto-resolved from meta$Platform
```

If the platform is **already known**, pass it directly:

```{r, eval = FALSE}
# runMethQC() loads IDAT, extracts beta + detection p-values, computes QC
qcset <- runMethQC(meta, platform = "EPIC")

# Summarize QC
qc_summary <- qc_summarize(qcset)
qc_summary$platform
qc_summary$n_samples
qc_summary$n_probes
```

Visualize QC metrics:

```{r, eval = FALSE}
plot(qcset, type = "qc_bar")                          # PASS/FAIL bar chart
plot(qcset, type = "intensity",  icutoff = 11)        # intensity scatter
plot(qcset, type = "detection_pval", pcutoff = 0.05)  # detection p-value per sample
plot(qcset, type = "probe_coverage")                  # % probes detected per sample
plot(qcset, type = "predicted_sex")                   # predicted sex distribution
plot(qcset, type = "ctrl_metrics")                    # ewastools scores (if present)
plot(qcset, type = "ctrl_metrics_detail",             # per-metric detail PDFs
     outFile = "qc_ctrl_metrics_detail")
```

Export QC-cleaned data:

```{r, eval = FALSE}
export(
  qcset,
  outdir = "qc_output",
  format = c("xlsx", "rds")
)
```

**If you already have QC-cleaned data:**

```{r, eval = FALSE}
# Data is already preprocessed - skip MethQcSet
meta <- LoadMeta("meta.tsv")
beta <- LoadBeta("cleaned_beta.tsv")  # Already QC-approved
```

## 2. Load Data and Run Function-Based Analysis

Load and prepare your data:

```{r, eval = FALSE}
input <- LoadMetaBeta(meta, beta, probeset = NULL)
meta <- input$meta
beta <- input$beta
```

## 3. Core Imprinting Analysis

Use `AnalyzeImprintStatus()` for sample-level IDS/Angle/mechanism output.

```{r, eval = FALSE}
res <- AnalyzeImprintStatus(
  betaFile = beta,
  metaFile = meta,
  probeset = "selected",
  ids_cutoff = 0.2
)

head(res)
```

Expected columns include:

- `paternal_median`
- `maternal_median`
- `IDS`
- `Angle`
- `Mechanism`
- `Status`
- `Confidence`

## 4. Visualize Results

### Polar Plot (IDS + Angle)

```{r, eval = FALSE}
p_polar <- PlotPolar(
  data = res,
  colorColumn = "SAMPLE_GROUP",
  title = "ImprintomeR:Polar",
  outFile = "polar_selected.pdf"
)
```

### Optional Additional Plots

```{r, eval = FALSE}
# Probe-level distributions
p_bee <- BetaBeePlot(beta, meta, SAMPLEID = "Sample_Name", outFile = "beeswarm.pdf")

# Mirror density by parental origin
p_mirror <- MirrorDensity(
  betaFile = beta,
  metaFile = meta,
  probeset = "selected",
  outFile = "mirror_density.pdf"
)
```

## 5. Use the ImprintomeSet API

`ImprintomeSet` stores inputs and derived outputs in one object.

```{r, eval = FALSE}
# probeset can be a data.frame or list with probe annotations
probesets <- readRDS(system.file("extdata", "probesets_hg19.rds", package = "imprintomeR"))
ps_selected <- probesets[["selected"]]

x <- ImprintomeSet(
  beta = beta,
  meta = meta,
  probeset = ps_selected,
  genome = "hg19",
  assay = "EPICv1"
)
```

Run the core analysis and store output in `results(x)`:

```{r, eval = FALSE}
x <- runImprintome(x, probeset = "selected", ids_cutoff = 0.2)
names(results(x))
```

Summarize object contents:

```{r, eval = FALSE}
s <- summarize(x)
s$object
s$results
s$plots
```

`Meth_QC()` returns a named list of core tabular QC artifacts. By default this
includes metadata and beta summaries, and it can also include detection
p-value summaries and minfi-derived intensity summaries when those inputs are
supplied.

Plot from stored plots or fallback to IDS/Angle results:

Available `plot_type` values in the `ImprintomeSet` dispatcher are:

- `auto`
- `polar`
- `mirror_density`
- `beeswarm`
- `beeswarm_origin`
- `beeswarm_chr`
- `violin`
- `heatmap`
- `circular_heatmap`
- `cor_heatmap`
- `rainfall`
- `radar`

```{r, eval = FALSE}
p <- plot(
  x,
  plot_type = "polar",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "SAMPLE_GROUP",
  outFile = "imprintomeset_polar.pdf"
)

p2 <- plot(
  x,
  plot_type = "beeswarm",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  outFile = "imprintomeset_beeswarm.pdf"
)

p_bee_origin <- plot(
  x,
  plot_type = "beeswarm_origin",
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  title = "Beeswarm Origin (Cohort)",
  outFile = "imprintomeset_beeswarm_origin.pdf"
)

sample_id1 <- colnames(beta(x))[2]
p_bee_chr <- plot(
  x,
  plot_type = "beeswarm_chr",
  probeset = "selected",
  sample_id = sample_id1,
  chr = "chr11",
  title = "Beeswarm by Chromosome (Origin Split)",
  outFile = "imprintomeset_beeswarm_chr.pdf"
)

outFile <- paste0("imprintomeset_quickstart_rainfall_", sample_id1, ".pdf")
p_rainfall <- plot(
  x,
  plot_type = "rainfall",
  sample_id = sample_id1,
  probeset = "classifier3",
  outFile = outFile
)

outFile <- paste0("imprintomeset_quickstart_radar_", sample_id1, ".pdf")
p_radar <- plot(
  x,
  plot_type = "radar",
  probeset = "selected",
  sample_id = sample_id1,
  title = "ImprintomeSet Radar",
  outFile = outFile
)

# Optional: persist plot objects and retrieve later
p_list <- plots(x)
p_list[["beeswarm.origin.cohort"]] <- p_bee_origin
p_list[["beeswarm.chr.sample1"]] <- p_bee_chr
plots(x) <- p_list

p_origin_from_object <- plots(x)[["beeswarm.origin.cohort"]]
p_chr_from_object <- plots(x)[["beeswarm.chr.sample1"]]
```

Export results (and optionally plots):

```{r, eval = FALSE}
manifest <- export(
  x,
  outdir = "imprintome_export",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

head(manifest)
```

At present, `export()` writes `results(x)` and optionally `plots(x)`.

## 6. Reproducibility Notes

- Keep metadata sample names aligned with beta matrix column names.
- Preserve probeset origin labels (`maternal`, `paternal`) for allele-aware metrics.
- Avoid changing IDS/Angle/IDI formulas unless the methods contract is updated.
- Always run MethQcSet QC first to ensure data quality.
- Document platform type and array version for traceability.

## Session Info

```{r, eval = FALSE}
sessionInfo()
```
