---
title: "imprintomeR Basic Workflow (Non-S4)"
author: "imprintomeR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{imprintomeR Basic Workflow (Non-S4)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates a complete `imprintomeR` workflow using core functions without relying on S4 objects such as `ImprintomeSet`.

### 1. Load Data

```{r, eval=FALSE}
library(imprintomeR)

meta <- LoadMeta("meta.tsv")
beta <- LoadBeta("cleaned_beta.tsv")
```

### 2. Run Analysis

```{r, eval=FALSE}
res <- AnalyzeImprintStatus(betaFile = beta, metaFile = meta, probeset = "selected", ids_cutoff = 0.2)
head(res)
```

### 3. Visualization

#### Polar Plot

```{r, eval=FALSE}
p_polar <- PlotPolar(
  data = res,
  colorColumn = "Sample_Group",
  title = "imprintomeR Polar Plot",
  outFile = "polar.pdf"
)
```

#### Beeswarm Plot

```{r, eval=FALSE}
p_beeswarm <- BetaBeePlot(
  beta = beta,
  meta = meta,
  SAMPLEID = "Sample_Name",
  probeset = "selected",
  outFile = "beeswarm.pdf"
)
```


#### Heatmap by Gene

You can generate gene-level aggregated methylation heatmaps directly using the `BetaHeatmapByGene()` function:

```r
hm <- BetaHeatmapByGene(
  beta = beta,
  meta = meta,
  probeset = "selected",
  SAMPLEID = "Sample_Name",
  annoColumn = "Sample_Group",
  clusterRows = TRUE,
  clusterColumns = TRUE,
  outFile = "heatmap_by_gene.pdf"
)
```

This function aggregates probe-level beta values by gene using the median within each gene and produces a hierarchical clustered heatmap useful for identifying imprinting patterns across the sample cohort.

## Session Info

```{r, eval=FALSE}
sessionInfo()
```
