---
title: "ImprintomeSet Workflow Quickstart"
author: "imprintomeR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ImprintomeSet Workflow Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette presents a Bioconductor-style, object-first workflow using imprintomeR's preprocessing-first architecture:

1. **MethQcSet**: Quality control for single-platform raw data
2. **ImprintomeSet**: Analysis container for imprinting studies

You will learn how to:

- create a `MethQcSet` for preprocessing
- run platform-specific QC
- convert to `ImprintomeSet` for analysis
- compute imprinting metrics with `runImprintome()`
- summarize, visualize, and export results

Core definitions used by imprintomeR:

- IDS: `sqrt((paternal_median - 0.5)^2 + (maternal_median - 0.5)^2)`
- Angle: derived from paternal/maternal deviations around `0.5`
- IDI: `(beta - 0.5) * 2`

## Package Startup

```{r, eval = FALSE}
library(imprintomeR)
```

## 1. QC Preprocessing (Optional if Data Already Cleaned)

**If you have RAW data:**

Load metadata. `meta.tsv` must be a tab-delimited file with at least `Sample_Name` and `Sample_Group` columns.

**For QC workflow (raw IDAT files):** also include `Basename` column (IDAT file path prefixes, e.g. `/path/to/idats/SentrixID_SentrixPosition`).

**For direct analysis (QC already done):** `Basename` is not needed.

```{r, eval = FALSE}
meta <- LoadMeta("meta.tsv")
```

If the platform is **unknown or mixed**, detect it first:

```{r, eval = FALSE}
# Reads one IDAT per sample, infers platform from probe count
# Appends Platform, Array_Size, Manifest, Status columns to meta
meta <- check_platform(meta)
table(meta$Platform)  # inspect; split if multiple platforms present

# Run QC for one platform at a time
meta_epic <- meta[meta$Platform == "EPIC", ]
qcset <- runMethQC(meta_epic)  # platform auto-resolved from meta$Platform
```

If the platform is **already known**, pass it directly:

```{r, eval = FALSE}
# runMethQC() loads IDAT files, extracts beta + detection p-values, and runs QC
qcset <- runMethQC(meta, platform = "EPIC")
```

Inspect QC results:

```{r, eval = FALSE}
qc_summary <- qc_summarize(qcset)
qc_summary$platform
qc_summary$n_samples
qc_summary$n_probes
qc_summary$aggregation_status
```

Visualize QC metrics:

```{r, eval = FALSE}
# PASS / FAIL sample count bar chart (default)
p_bar <- plot(qcset, type = "qc_bar")

# Methylated vs unmethylated intensity scatter, cutoff line at log2 = 11
p_int <- plot(qcset, type = "intensity", icutoff = 11,
              outFile = "qc_intensity.pdf")

# Average detection p-value per sample, cutoff line at 0.05
p_dp  <- plot(qcset, type = "detection_pval", pcutoff = 0.05)

# % probes detected per sample, reference line at 95%
p_cov <- plot(qcset, type = "probe_coverage")

# Predicted sex distribution
p_sex <- plot(qcset, type = "predicted_sex")

# ewastools control probe metrics — summary (returns NULL silently if absent)
p_ctrl <- plot(qcset, type = "ctrl_metrics")

# Per-metric detail plots (Bisulfite Conversion, Specificity, Non-polymorphic)
# outFile is used as a path prefix; one PDF is saved per control metric
plot(qcset, type = "ctrl_metrics_detail",
     outFile = "qc_ctrl_matrix_detail")
```

Export cleaned data:

```{r, eval = FALSE}
dir.create("qc_output", showWarnings = FALSE)

exported_files <- export(
  qcset,
  outdir = "qc_output",
  format = c("xlsx", "rds")
)
# Creates two Excel workbooks:
#   qc_table_main.xlsx: metadata, QC_matrix, Ctrl_matrix (if available), statistics
#   qc_table_extra.xlsx: QC_metrics (thresholds), recall_rate (if available), 
#                         contamination (if available), predUniqDonor_ID (if available)
# Plus: qcset.rds (full MethQcSet object)
# Plus: qcset_summary.txt (summary report with Notes section documenting missing sheets)
```

**If you already have QC-cleaned data:**

Skip MethQcSet and load pre-cleaned beta values. Note: `Basename` column is not needed since QC is already done.

```{r, eval = FALSE}
meta <- LoadMeta("meta.tsv")  # Only needs Sample_Name and Sample_Group
beta <- LoadBeta("cleaned_beta.tsv")  # Already QC-approved

# Proceed to Section 2: Create ImprintomeSet
```

Optional: Merge multiple QC-approved single-platform datasets:

```{r, eval = FALSE}
# Two separate MethQcSet objects (same platform)
qcset1 <- MethQcSet(meta1, beta1, platform = "EPICv1")
qcset2 <- MethQcSet(meta2, beta2, platform = "EPICv1")

# Synchronized merge (common samples/probes)
merged <- merge(qcset1, qcset2, how = "inner")

# Find common elements across objects
common_samples <- find_intersection(qcset1, qcset2, by = "samples")
common_probes <- find_intersection(qcset1, qcset2, by = "probes")
```

## 2. Create ImprintomeSet from QC-Approved Data

Once your data passes QC, convert to analysis container:

```{r, eval = FALSE}
# Load probeset annotations
probesets <- readRDS(
  system.file("extdata", "probesets_hg19.rds", package = "imprintomeR")
)

# EPICv2 only: aggregate replicate probes before conversion (required)
# Reduces ~937k probes to ~865k unique base probes (cg... IDs)
# as.ImprintomeSet() will error if this step is skipped for EPICv2.
if (qcset@platform == "EPICv2") {
  qcset <- aggregate_probes(qcset)   # idempotent — safe to call again
}

# Convert MethQcSet to ImprintomeSet
x <- as.ImprintomeSet(
  qcset,
  probeset = probesets[["selected"]],
  genome = "hg19"
)
```

Alternatively, construct `ImprintomeSet` directly if you've already completed QC:

```{r, eval = FALSE}
x <- ImprintomeSet(
  beta = beta,
  meta = meta,
  probeset = probesets[["selected"]],
  genome = "hg19",
  assay = "EPICv1"
)
```

Constructor validation includes:

- overlapping sample IDs between `meta$Sample_Name` and beta columns
- required metadata columns (`Sample_Name`, `Sample_Group`)
- non-empty beta/probeset objects

The object keeps derived outputs in separate slots:

- `results(x)` for imprinting analysis tables
- `plots(x)` for stored plot objects

## 3. Run Core Analysis with runImprintome()

`runImprintome()` stores the analysis table inside `results(x)`.

```{r, eval = FALSE}
x <- runImprintome(
  x,
  probeset = "selected",
  ids_cutoff = 0.2
)
```

You can run additional probesets and keep all outputs in the same object.

```{r, eval = FALSE}
x <- runImprintome(
  x,
  probeset = "classifier3",
  ids_cutoff = 0.2,
  result_name = "AnalyzeImprintStatus.classifier3"
)
```

## 4. Inspect Results

```{r, eval = FALSE}
result_names <- names(results(x))
result_names

res <- results(x)[["AnalyzeImprintStatus.selected"]]
head(res)
```

Expected columns include:

- `paternal_median`
- `maternal_median`
- `IDS`
- `Angle`
- `Mechanism`
- `Status`
- `Confidence`

## 5. Summarize Object State

```{r, eval = FALSE}
s <- summarize(x)
s$object
s$results
s$plots
```

## 6. Access and Update Slots

```{r, eval = FALSE}
# Accessors
head(beta(x))
head(meta(x))
assay(x)
genome(x)

# Replacement examples
assay(x) <- "EPICv2"
genome(x) <- "hg38"
```

## 7. Plot with ImprintomeSet Dispatcher

Use `plot_type` to choose a visualization backend while keeping one unified
object-first interface.

Supported `plot_type` values:

- `auto`
- `polar`
- `mirror_density`
- `beeswarm`
- `beeswarm_origin`
- `beeswarm_chr`
- `violin`
- `heatmap`
- `heatmap_by_gene`
- `circular_heatmap`
- `cor_heatmap`
- `rainfall`
- `radar`

### 7.1 Example Code for Every Plot Type

For a compact one-block demo of all plot backends, see
*ImprintomeSet Visualization Gallery*, Section 2.3.

```{r, eval = FALSE}
# Use a real sample identifier for sample-level views
sample_id1 <- colnames(beta(x))[2]

# 1) auto: returns most recent stored plot when available, else polar fallback
p_auto <- plot(
  x,
  plot_type = "auto",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "SAMPLE_GROUP",
  title = "ImprintomeSet Auto",
  outFile = "imprintomeset_quickstart_auto.pdf"
)

# 2) polar
p_polar <- plot(
  x,
  plot_type = "polar",
  result_name = "AnalyzeImprintStatus.selected",
  colorColumn = "SAMPLE_GROUP",
  title = "ImprintomeSet Polar",
  outFile = "imprintomeset_quickstart_polar.pdf"
)

# 3) mirror_density
p_mirror <- plot(
  x,
  plot_type = "mirror_density",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  outFile = "imprintomeset_quickstart_mirror.pdf"
)

# 4) beeswarm
p_bee <- plot(
  x,
  plot_type = "beeswarm",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  outFile = "imprintomeset_quickstart_beeswarm.pdf"
)

# 5) beeswarm_origin
p_bee_origin <- plot(
  x,
  plot_type = "beeswarm_origin",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  title = "Beeswarm Origin (Cohort)",
  outFile = "imprintomeset_quickstart_beeswarm_origin.pdf"
)

# 6) beeswarm_chr
p_bee_chr <- plot(
  x,
  plot_type = "beeswarm_chr",
  probeset = "selected",
  sample_id = sample_id1,
  chr = "chr11",
  title = "Beeswarm by Chromosome (Origin Split)",
  outFile = "imprintomeset_quickstart_beeswarm_chr.pdf"
)

# 7) violin
p_violin <- plot(
  x,
  plot_type = "violin",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  outFile = "imprintomeset_quickstart_violin.pdf"
)

# 8) heatmap_by_probe
p_heatmap <- plot(
  x,
  plot_type = "heatmap_by_probe",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  outFile = "imprintomeset_quickstart_heatmap.pdf"
)

# 9) heatmap_by_gene (aggregates probes to gene level)
p_heatmap_by_gene <- plot(
  x,
  plot_type = "heatmap_by_gene",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  annoColumn = "SAMPLE_GROUP",
  clusterRows = TRUE,
  clusterColumns = TRUE,
  outFile = "imprintomeset_quickstart_heatmap_by_gene.pdf"
)

# 10) circular_heatmap
p_circular <- plot(
  x,
  plot_type = "circular_heatmap",
  SAMPLEID = "SAMPLE_NAME",
  sectionColumn = "SAMPLE_GROUP",
  probeset = "selected",
  outFile = "imprintomeset_quickstart_circular_heatmap.pdf"
)

# 10) cor_heatmap
cor_mat <- plot(
  x,
  plot_type = "cor_heatmap",
  probeset = "selected",
  SAMPLEID = "SAMPLE_NAME",
  prefix = "imprintomeset_quickstart"
)

# 12) rainfall
outFile <- paste0("imprintomeset_quickstart_rainfall_", sample_id1, ".pdf")
p_rainfall <- plot(
  x,
  plot_type = "rainfall",
  sample_id = sample_id1,
  probeset = "classifier3",
  outFile = outFile
)

# 13) radar
outFile <- paste0("imprintomeset_quickstart_radar_", sample_id1, ".pdf")
p_radar <- plot(
  x,
  plot_type = "radar",
  probeset = "selected",
  sample_id = sample_id1,
  title = "ImprintomeSet Radar",
  outFile = outFile
)
```

`beeswarm_chr` returns a chromosome-faceted plot for one sample with
maternal/paternal split points and per-group median summary markers.

### 7.2 Access Plot Objects Stored in ImprintomeSet

```{r, eval = FALSE}
# Save selected plot outputs into the object
p_list <- plots(x)
p_list[["auto.selected"]] <- p_auto
p_list[["polar.selected"]] <- p_polar
p_list[["mirror.selected"]] <- p_mirror
p_list[["beeswarm.selected"]] <- p_bee
p_list[["beeswarm_origin.selected"]] <- p_bee_origin
p_list[["beeswarm_chr.sample1"]] <- p_bee_chr
p_list[["violin.selected"]] <- p_violin
p_list[["heatmap.selected"]] <- p_heatmap
p_list[["heatmap_by_gene.selected"]] <- p_heatmap_by_gene
p_list[["circular.selected"]] <- p_circular
p_list[["rainfall.sample1"]] <- p_rainfall
p_list[["radar.sample1"]] <- p_radar
plots(x) <- p_list

# List available stored plots
names(plots(x))

# Retrieve one plot object by name
p_from_object <- plots(x)[["beeswarm_origin.selected"]]
p_from_object

# Retrieve a chromosome-faceted beeswarm plot
p_chr_from_object <- plots(x)[["beeswarm_chr.sample1"]]
p_chr_from_object
```

## 8. Export Results

```{r, eval = FALSE}
manifest <- export(
  x,
  outdir = "imprintome_results",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

head(manifest)
```

`export()` writes `results(x)` and optionally `plots(x)`.

**For QC preprocessing (MethQcSet)**, the export includes `summary.txt` with a "Notes" section documenting any optional sheets not created due to missing underlying data. Check it to understand data availability:

```{r, eval = FALSE}
# View what sheets were and were not created
readLines("qc_output/methqcset_summary.txt")
```

## 9. Reproducibility Notes

- Keep beta columns aligned with `meta$SAMPLE_NAME`.
- Preserve maternal/paternal origin labels in probeset annotations.
- Keep metric definitions unchanged unless method contracts are explicitly updated.
- For reproducible QC workflows, save MethQcSet summaries with `qc_summarize()`.
- Document platform and array type for reference.

## Session Info

```{r, eval = FALSE}
sessionInfo()
```
