---
title: "WGBS Region-Level Imprinting Workflow"
subtitle: "Using wgbstools beta_to_table output with Rosenski_region"
author: "imprintomeR contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{WGBS Region-Level Imprinting Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "WGBS_region_",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This vignette demonstrates the region-level WGBS workflow added for Rosenski-style imprinting regions. It starts from a `wgbstools beta_to_table` output table rather than Illumina array beta values.

You will:

1. Load a region-level beta table from `wgbstools beta_to_table`
2. Align regions to `probesets_<genome>.rds[["Rosenski_region"]]`
3. Create a WGBS-backed `ImprintomeSet`
4. Run `runImprintome()` with `probeset = "Rosenski_region"`
5. Generate the region-safe default visualizations

This workflow is intended for **aggregated region-level methylation values**. For CpG-level WGBS extraction, see the `wgbstools_beta_subset.py` helper workflow.

---

## Phase 1: Prepare wgbstools Region Beta Table

The input is the standard `wgbstools beta_to_table` table. Its first columns are genomic interval coordinates and CpG-index bounds; remaining columns are samples.

Example command outside R:

```{bash, eval = FALSE}
wgbstools beta_to_table \
  Rosenski_refined_iDMRs_hg19.withCpG.bed \
  --betas *.beta \
  --output Rosenski_iDMRs_mean_beta.hg19.tsv
```

After running `wgbstools beta_to_table`, the generated table has this shape:

```{r preview-input, eval = FALSE}
library(imprintomeR)

beta_file <- "Rosenski_iDMRs_mean_beta.hg19.tsv"
read.delim(beta_file, check.names = FALSE, nrows = 3)
```

Expected columns:

```text
chr  start  end  startCpG  endCpG  sample1  sample2  ...
```


---

## Phase 2: Load WGBS Region Beta Values

`LoadWGBSRegionBeta()` converts each interval to a stable region ID:

```text
NAME = chr:start-end
```

By default, it aligns the rows to `probesets_hg19.rds[["Rosenski_region"]]`; use `genome = "hg38"` for hg38 region tables.

```{r load-region-beta, eval = FALSE}
beta_all <- LoadWGBSRegionBeta(beta_file, genome = "hg19")

dim(beta_all)
head(rownames(beta_all))
colnames(beta_all)

# Region/probeset annotations are retained as attributes
head(attr(beta_all, "regions"))
head(attr(beta_all, "probeset"))
```

For the refined `Rosenski_region` signature, generate the region-level beta table from `Rosenski_refined_iDMRs_hg19.withCpG.bed` so the input coordinates match `probesets_hg19.rds[["Rosenski_region"]]`.

If you want to keep all regions exactly as they appear in the input file, set `probeset = NULL`:

```{r load-no-alignment, eval = FALSE}
beta_all <- LoadWGBSRegionBeta(beta_file, probeset = NULL)
```

---

## Phase 3: Create Metadata

For direct WGBS analysis, metadata only needs to match the sample columns in the beta matrix. `Sample_Name` must match `colnames(beta_region)`.

```{r create-meta, eval = FALSE}
sample_ids <- colnames(beta_all)

sample_group <- ifelse(
  grepl("-cfDNA-", sample_ids, ignore.case = TRUE),
  "cfDNA",
  ifelse(
    grepl("-WBC-", sample_ids, ignore.case = TRUE),
    "WBC",
    sub("-Z[0-9A-Za-z]+$", "",
        sub("^GSM[0-9]+_", "", sample_ids))
  )
)
meta_region <- data.frame(
  Sample_Name = colnames(beta_region),
  Sample_Group = sample_group,
  stringsAsFactors = FALSE
)

meta_region
```

---

## Phase 4: Create an ImprintomeSet

Region-level WGBS objects should explicitly set `genome = "hg19"` and `assay = "WGBS"`.

```{r create-imprintomeset, eval = FALSE}
probesets <- readRDS(system.file(
  "extdata",
  "probesets_hg19.rds",
  package = "imprintomeR"
))

x <- ImprintomeSet(
  beta = beta_region,
  meta = meta_region,
  probeset = probesets[["Rosenski_region"]],
  genome = "hg19",
  assay = "WGBS"
)

x
```

---

## Phase 5: Run Region-Level Imprinting Analysis

`runImprintome()` uses the maternal and paternal `ORIGIN` annotations in `Rosenski_region` to calculate paternal/maternal medians, IDS, Angle, and imprinting status.

```{r run-imprintome, eval = FALSE}
x <- runImprintome(
  x,
  probeset = "Rosenski_region",
  ids_cutoff = 0.2
)

status <- results(x)[["AnalyzeImprintStatus.Rosenski_region"]]
status[, c(
  "Sample_Name",
  "paternal_median",
  "maternal_median",
  "IDS",
  "Angle",
  "Mechanism",
  "Status"
)]
```

Interpretation should remain cautious because WGBS region-level beta values are not allele-specific. IDS and Angle summarize directional deviation from balanced methylation across maternal-origin and paternal-origin regions.

---

## Phase 6: Region-Safe Default Visualizations

For `probeset = "Rosenski_region"`, the default visualization set is intentionally restricted to plot types that work with region-level beta rows:

```text
polar
beeswarm_origin
heatmap_by_gene
radar
```

```{r run-region-visualizations, eval = FALSE}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  result_name = "AnalyzeImprintStatus.Rosenski_region",
  plot_types = "default",
  save_plots = FALSE,
  store_plots = TRUE
)

attr(x, "visualization_manifest")
names(plots(x))
```

### Individual Plot Examples

```{r plot-polar, eval = FALSE, fig.width = 8, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  result_name = "AnalyzeImprintStatus.Rosenski_region",
  plot_types = "polar",
  save_plots = FALSE,
  store_plots = TRUE
)
print(plots(x)[["polar.Rosenski_region"]])
```

```{r plot-beeswarm-origin, eval = FALSE, fig.width = 8, fig.height = 5}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  plot_types = "beeswarm_origin",
  save_plots = FALSE,
  store_plots = TRUE
)
print(plots(x)[["beeswarm_origin.Rosenski_region"]])
```

```{r plot-heatmap-gene, eval = FALSE, fig.width = 8, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  plot_types = "heatmap_by_gene",
  save_plots = FALSE,
  store_plots = TRUE
)
plots(x)[["heatmap_by_gene.Rosenski_region"]]
```

```{r plot-radar, eval = FALSE, fig.width = 8, fig.height = 8}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  result_name = "AnalyzeImprintStatus.Rosenski_region",
  plot_types = "radar",
  save_plots = FALSE,
  store_plots = TRUE
)
print(plots(x)[[grep("^radar\\.", names(plots(x)), value = TRUE)[1]]])
```

Avoid using `plot_types = "all"` for `Rosenski_region` unless you explicitly want to test non-region-aware plot paths. In particular, classic `beeswarm` and direct `rainfall` are probe-oriented and are not part of the region-safe default workflow.

---

## Phase 7: Export Results

After analysis and optional visualizations, export result tables and plot files.

```{r export-results, eval = FALSE}
manifest <- export(
  x,
  outdir = "Rosenski_region_imprintome_output",
  prefix = "Rosenski_iDMRs",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

manifest
```

The export will include available result tables such as:

```text
Rosenski_iDMRs_results_AnalyzeImprintStatus.Rosenski_region.tsv
Rosenski_iDMRs_imprintomeSet.rds
```

## Toy Example: Submit WGBS Summary and Run imprintomeR CLI

The following example summarizes WGBS `.beta` files over the refined Rosenski iDMR regions, submits the job to LSF, runs the imprintomeR CLI on the region-level beta table, and then exports a radar plot for one selected sample.

```{bash summarize-wgbs-regions, eval = FALSE}
# Summarize methylation over refined Rosenski iDMR regions
cmd="wgbstools beta_to_table --betas *.beta \
--output Rosenski_iDMRs_mean_beta.hg19.tsv \
Rosenski_refined_iDMRs_hg19.withCpG.bed"

prefix=wgbstools_iDMRs_beta
bsub -q priority -J ${prefix} -M 50000 -oo ${prefix}.log -eo ${prefix}.errlog "$cmd" 2>&1 >/dev/null
```

```{bash run-wgbs-cli, eval = FALSE}
# Run the imprintomeR WGBS region workflow
Rscript /PATH/TO/imprintomeR/scripts/run_imprintomeR.R \
  -B Rosenski_iDMRs_mean_beta.hg19.tsv \
  -m GSE186458_meta.txt \
  -o wgbs_imprintome_results_mean \
  --prefix Rosenski_WGBS \
  --probeset Rosenski_region \
  --genome hg19 \
  --plot-types default \
  -v
```

```{r radar-specific-sample, eval = FALSE, fig.width = 8, fig.height = 8}
# Generate a radar plot for one specific sample
library(imprintomeR)

sample_id1 <- "GSM5652245_Pancreas-Acinar-Z0000043X"
x <- readRDS("wgbs_imprintome_results_mean/Rosenski_WGBS_imprintomeSet.rds")

x <- runImprintomeVisualizations(
  x,
  outdir = "wgbs_imprintome_results_mean",
  plot_types = "radar",
  probeset = "Rosenski_region",
  sample_id = sample_id1,
  prefix = "Rosenski_WGBS",
  store_plots = FALSE,
  save_plots = TRUE
)
```
---

## Notes and Limitations

- `Rosenski_region` is a **region-level** signature, not an Illumina probe-level probeset.
- `ORIGIN` is derived from primary germline DMR evidence in Rosenski et al. Sheet S10.
- `Oocyte gDMR` is treated as maternal; `Sperm gDMR` is treated as paternal.
- Secondary-only or no-primary-germline regions have `ORIGIN = NA` and do not drive maternal/paternal scoring.
- Region-level WGBS beta values are not allele-specific; interpretation should be cautious and supported by orthogonal evidence when possible.
