---
title: "WGBS and EM-seq Region-Level Imprinting Workflow"
subtitle: "Using wgbstools, ONT bedMethyl, WGBS, or EM-seq region beta output with Rosenski_region"
author: "imprintomeR contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{WGBS Region-Level Imprinting Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "WGBS_region_",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This vignette demonstrates the region-level WGBS/ONT workflow added for Rosenski-style imprinting regions. It starts from region-level beta tables generated from `wgbstools beta_to_table`, ONT/modkit bedMethyl, short-read WGBS, or EM-seq CpG bedGraph files, rather than Illumina array beta values.

You will:

1. Generate or load a region-level beta table from `wgbstools beta_to_table`, ONT/modkit bedMethyl, short-read WGBS, or EM-seq CpG bedGraph files
2. Align regions to `probesets_<genome>.rds[["Rosenski_region"]]` or the focused `chr11p15_region` subset
3. Create a WGBS-backed `ImprintomeSet`
4. Run `runImprintome()` with `probeset = "Rosenski_region"` or `probeset = "chr11p15_region"`
5. Generate the region-safe default visualizations

This workflow is intended for **aggregated region-level methylation values**. For CpG-level WGBS extraction, see the `wgbstools_beta_subset.py` helper workflow. For ONT/modkit bedMethyl files, use `inst/scripts/parse_ONT_bedMethyl.R` to create the same region-level beta table format.

---

## Phase 1A: Prepare wgbstools Region Beta Table

The input is the standard `wgbstools beta_to_table` table. Its first columns are genomic interval coordinates and CpG-index bounds; remaining columns are samples. Install and configure [wgbstools](https://github.com/nloyfer/wgbs_tools) following the package instructions, including genome initialization for the build you plan to use. imprintomeR ships the canonical region BED (`Rosenski_refined_iDMRs_hg19.bed`); the `.withCpG.bed` file is a generated wgbstools intermediate and does not need to be stored in `inst/extdata`.

Example command outside R:

```{bash, eval = FALSE}
# Generate the CpG-index BED intermediate from the canonical package BED.
wgbstools convert \
  inst/extdata/Rosenski_refined_iDMRs_hg19.bed \
  --genome hg19 \
  -L \
  -o Rosenski_refined_iDMRs_hg19.withCpG.bed

wgbstools beta_to_table \
  --betas *.beta \
  --output Rosenski_iDMRs_mean_beta.hg19.tsv \
  Rosenski_refined_iDMRs_hg19.withCpG.bed
```

After running `wgbstools beta_to_table`, the generated table has this shape:

```{r preview-input, eval = FALSE}
library(imprintomeR)

beta_file <- "Rosenski_iDMRs_mean_beta.hg19.tsv"
read.delim(beta_file, check.names = FALSE, nrows = 3)
```

Expected columns:

```text
chr  start  end  startCpG  endCpG  sample1  sample2  ...
```

---

## Phase 1B: Parse ONT BedMethyl Files

For Oxford Nanopore/modkit bedMethyl-style outputs, use `parse_ONT_bedMethyl.R` to summarize CpG-level 5mC calls over the hg38 Rosenski refined iDMR regions and create an imprintomeR-compatible region beta table.

The parser:

- Intersects each bedMethyl file with `Rosenski_refined_iDMRs_hg38.bed`.
- Keeps only modified-base code `m`, corresponding to 5mC.
- Combines modified and valid coverage counts across phased files belonging to the same sample.
- Writes intermediate sample-region counts and a final beta matrix for `run_imprintomeR.R -B`.

Input bedMethyl files should follow the common modkit/bedMethyl-style layout:

```text
chr1  10468  10469  m  80  .  10468  10469  255,0,0  80  21.25  17  63  0  2  911  11
chr1  10468  10469  h  80  .  10468  10469  255,0,0  80   0.00   0  17  63 2  911  11
```

The default parser columns are:

- column 1: chromosome
- columns 2-3: 0-based genomic interval
- column 4: modified-base code; only `m` is kept by default
- column 10: valid coverage, used as `N_valid_cov`
- column 12: modified count, used as `N_mod`

Rows with other modified-base codes, such as `h`, are ignored unless `--mod-code` is changed.

Example command for files one directory below the current directory:

```{bash parse-ont-bedmethyl, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/parse_ONT_bedMethyl.R \
  --bedmethyl "./*/*.bedmethyl.gz" \
  --regions /PATH/TO/imprintomeR/extdata/Rosenski_refined_iDMRs_hg38.bed \
  --outdir ont_bedmethyl_region_beta \
  --prefix ONT_Rosenski_hg38 \
  --jobs 8 \
  -v
```

`"./*/*.bedmethyl.gz"` matches phased files such as:

```text
sample1/sample1.HP1.bedmethyl.gz
sample1/sample1.HP2.bedmethyl.gz
sample2/sample2.HP1.bedmethyl.gz
sample2/sample2.HP2.bedmethyl.gz
```

It also matches non-phased files when each sample has its own directory:

```text
sample1/mods.bedmethyl.gz
sample2/mods.bedmethyl.gz
```

For generic filenames, non-phased files such as `mods.bedmethyl.gz` or `wf_mods.bedmethyl.gz` use the parent directory as `Sample_Name`. Phased file pairs such as `mods.1.bedmethyl.gz` / `mods.2.bedmethyl.gz` or `wf_mods.1.bedmethyl.gz` / `wf_mods.2.bedmethyl.gz` also use the parent directory as `Sample_Name`, so counts are combined for the same sample. Do not use `"./*/.bedmethyl.gz"` unless the files are literally named `.bedmethyl.gz`.

Main outputs:

- `{prefix}_sample_map.tsv`: maps input files to inferred `Sample_Name` values.
- `{prefix}_file_region_counts.tsv`: one row per input file and region before phase/sample merging.
- `{prefix}_sample_region_counts.tsv`: combined counts after files with the same `Sample_Name` are merged.
- `{prefix}_beta.tsv`: final imprintomeR-compatible region beta table.

The final beta table has coordinate columns followed by sample beta columns:

```text
chr  start  end  sample1  sample2  sample3
chr11  2016392  2022700  0.519342  0.486019  0.827114
chr11  2721224  2729557  0.504831  0.492771  0.218806
```

For ONT bedMethyl, region beta is calculated as:

```text
sum(N_mod) / sum(N_valid_cov)
```

Missing or non-numeric count values are coerced to `0` by the AWK parser. If a row has valid coverage but a missing or non-numeric methylated count, beta may be biased downward.

Run the imprinting workflow with the generated hg38 beta table:

```{bash run-ont-cli, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/run_imprintomeR.R \
  -B ont_bedmethyl_region_beta/ONT_Rosenski_hg38_beta.tsv \
  -m ONT_meta.txt \
  -o ONT_imprintome_results \
  --prefix ONT_Rosenski_hg38 \
  --probeset Rosenski_region \
  --genome hg38 \
  --plot-types default \
  -v
```

For phased files with names that do not collapse cleanly to one sample ID, provide `--sample-map` with columns `file` and `Sample_Name`. For large cohorts, use `--jobs` to process multiple bedMethyl files in parallel on Linux/HPC; start with `--jobs 8` or `--jobs 12`, then adjust based on queue limits, I/O performance, and available CPU cores.

---

## Phase 1C: Parse Short-Read WGBS Calls To Region Beta

For short-read WGBS methylation call files, use `parse_WGBS_to_region_beta.R` to summarize CpG-level calls over Rosenski refined iDMR regions and create the same imprintomeR-compatible region beta table used by the ONT workflow.

Supported input formats:

- `bismark_coverage`: Bismark coverage/bedGraph-style files. The parser uses columns 1-3 as `chr/start/end`, column 5 as methylated count, and column 6 as unmethylated count.

```text
chr1  10468  10469  66.6667  8  4
```

- `bismark_cytosine`: Bismark cytosine report files. The parser converts the 1-based cytosine position to a BED interval, keeps the requested context (`CG` by default), and uses columns 4-5 as methylated/unmethylated counts.

```text
chr1  10469  +  8  4  CG  CGA
```

- `wgbstools_beta2bed`: text BED output generated from `wgbstools beta2bed`. The parser uses columns 1-3 as `chr/start/end`, column 4 as methylated count, and column 5 as total coverage.

```text
chr1  10468  10470  8  12
```

- `bed_beta`: generic BED-like CpG methylation table. Two layouts are supported: columns 4-5 as methylated count and total coverage, or column 4 as beta when no count columns are available. Count-based input is preferred.

```text
chr1  10468  10469  8  12
chr1  10468  10469  0.6667
```

Example for Bismark coverage files:

```{bash parse-bismark-coverage, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/parse_WGBS_to_region_beta.R \
  --input "*.cov.gz" \
  --format bismark_coverage \
  --genome hg38 \
  --outdir wgbs_region_beta \
  --prefix WGBS_Rosenski_hg38 \
  --jobs 8 \
  -v
```

Example for Bismark cytosine reports:

```{bash parse-bismark-cytosine, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/parse_WGBS_to_region_beta.R \
  --input "*.CX_report.txt.gz" \
  --format bismark_cytosine \
  --context CG \
  --genome hg38 \
  --outdir wgbs_region_beta \
  --prefix WGBS_CG_Rosenski_hg38 \
  --jobs 8 \
  -v
```

Example for `wgbstools beta2bed` text output:

```{bash parse-wgbstools-beta2bed, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/parse_WGBS_to_region_beta.R \
  --input "*.beta.bed.gz" \
  --format wgbstools_beta2bed \
  --regions /PATH/TO/imprintomeR/extdata/Rosenski_refined_iDMRs_hg19.bed \
  --outdir wgbs_region_beta \
  --prefix WGBS_Rosenski_hg19 \
  --jobs 8 \
  -v
```

Main outputs:

- `{prefix}_sample_map.tsv`: inferred file-to-sample mapping.
- `{prefix}_file_region_counts.tsv`: per-file region counts.
- `{prefix}_sample_region_counts.tsv`: combined sample-region methylated counts, total counts, beta values, and site counts.
- `{prefix}_beta.tsv`: final imprintomeR-compatible beta table with `chr`, `start`, `end`, and sample columns.

The parser calculates region beta as:

```text
sum(N_methylated) / sum(N_total)
```

This count-weighted estimate is preferred over averaging CpG beta values when CpG coverage varies across a region. Missing or non-numeric count values are coerced to `0` by the AWK parser. If a row has valid coverage but a missing or non-numeric methylated count, beta may be biased downward.

Run imprintomeR with the generated beta table:

```{bash run-short-read-wgbs-cli, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/run_imprintomeR.R \
  -B wgbs_region_beta/WGBS_Rosenski_hg38_beta.tsv \
  -m WGBS_meta.txt \
  -o WGBS_imprintome_results \
  --prefix WGBS_Rosenski_hg38 \
  --probeset Rosenski_region \
  --genome hg38 \
  --plot-types default \
  -v
```

Use `--sample-map` with columns `file` and `Sample_Name` when filenames do not cleanly encode sample IDs.

---

## Phase 1D: Parse EM-seq CpG BedGraph Files

For EM-seq output summarized by Bismark, use
`parse_EMseq_to_region_beta.R` to process multiple CpG bedGraph files and
create an imprintomeR-compatible region beta table. The parser follows the
same region aggregation and output conventions as
`parse_WGBS_to_region_beta.R`, but is specialized for the six-column
EM-seq CpG bedGraph layout.

Expected input:

```text
track type="bedGraph" description="sample CpG merged methylation levels"
chr1  10468  10470  66  2  1
chr1  10470  10472  75  3  1
```

The columns are:

1. chromosome
2. BED start
3. BED end
4. methylation percentage; retained for reference but not used in aggregation
5. methylated count
6. unmethylated count

The parser skips `track`, `browser`, comment, and column-header lines. It
filters sites using `--min-site-coverage`, intersects retained CpGs with the
requested region BED, and calculates count-weighted region beta as:

```text
sum(methylated) / sum(methylated + unmethylated)
```

This is preferable to averaging the percentage column when CpG coverage varies
within a region.

Example for multiple uncompressed or gzipped bedGraph files:

```{bash parse-emseq-bedgraph, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/parse_EMseq_to_region_beta.R \
  --input "dev_tmp/BismarkRemapping/*_CpG.bedGraph" \
  --genome hg38 \
  --outdir EMseq_region_beta \
  --prefix EMseq_Rosenski_hg38 \
  --jobs 8 \
  -v
```

Multiple glob patterns can be comma-separated:

```{bash parse-emseq-multiple-patterns, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/parse_EMseq_to_region_beta.R \
  --input "batch1/*_CpG.bedGraph,batch2/*_CpG.bedGraph.gz" \
  --regions /PATH/TO/imprintomeR/extdata/Rosenski_refined_iDMRs_hg38.bed \
  --outdir EMseq_region_beta \
  --prefix EMseq_Rosenski_hg38 \
  --jobs 8
```

By default, filenames such as
`3197296_SJGENLK084660_G1-CCSS-32615701_CpG.bedGraph` produce the sample
name `3197296_SJGENLK084660_G1-CCSS-32615701`. When filenames do not encode
the desired sample IDs, provide `--sample-map` as a tab-delimited file with
columns `file` and `Sample_Name`. Multiple files assigned to the same
sample are combined by summing methylated and total counts before beta is
calculated.

Main outputs:

- `{prefix}_sample_map.tsv`: input-file to sample-name mapping.
- `{prefix}_file_region_counts.tsv`: per-file regional methylated and total counts.
- `{prefix}_sample_region_counts.tsv`: counts and beta after files are combined by sample.
- `{prefix}_beta.tsv`: imprintomeR-compatible matrix with `chr`, `start`,
  `end`, and one beta column per sample.

The final beta table can be supplied directly to the imprintomeR CLI:

```{bash run-emseq-cli, eval = FALSE}
Rscript /PATH/TO/imprintomeR/scripts/run_imprintomeR.R \
  -B EMseq_region_beta/EMseq_Rosenski_hg38_beta.tsv \
  -m EMseq_meta.tsv \
  -o EMseq_imprintome_results \
  --prefix EMseq_Rosenski_hg38 \
  --probeset Rosenski_region \
  --genome hg38 \
  --plot-types default \
  -v
```

The parser requires `bash`, `awk`, and `bedtools` and is intended for
Linux/HPC execution. Use `--jobs` to process independent bedGraph files in
parallel.

---

## Phase 2: Load Region Beta Values

`LoadWGBSRegionBeta()` converts each interval to a stable region ID:

```text
NAME = chr:start-end
```

By default, it aligns the rows to `probesets_hg19.rds[["Rosenski_region"]]`; use `genome = "hg38"` for hg38 region tables, including ONT bedMethyl summaries produced with `Rosenski_refined_iDMRs_hg38.bed`. For focused 11p15 analysis, use `probeset = "chr11p15_region"` with the matching genome build; this subset contains chr11 refined iDMRs and excludes OSBPL5.

```{r load-region-beta, eval = FALSE}
beta_all <- LoadWGBSRegionBeta(beta_file, genome = "hg19")

dim(beta_all)
head(rownames(beta_all))
colnames(beta_all)

# Region/probeset annotations are retained as attributes
head(attr(beta_all, "regions"))
head(attr(beta_all, "probeset"))
```

For the refined `Rosenski_region` signature, generate `Rosenski_refined_iDMRs_hg19.withCpG.bed` from the canonical `Rosenski_refined_iDMRs_hg19.bed`, then create the region-level beta table from that intermediate so the input coordinates match `probesets_hg19.rds[["Rosenski_region"]]`.

If you want to keep all regions exactly as they appear in the input file, set `probeset = NULL`:

```{r load-no-alignment, eval = FALSE}
beta_all <- LoadWGBSRegionBeta(beta_file, probeset = NULL)
```

If you want to focus on the chr11p15 region subset, use `chr11p15_region`:

```{r load-chr11p15-region, eval = FALSE}
beta_chr11p15 <- LoadWGBSRegionBeta(
  beta_file,
  probeset = "chr11p15_region",
  genome = "hg38"
)

nrow(beta_chr11p15)
rownames(beta_chr11p15)
```

---

## Phase 3: Create Metadata

For direct WGBS/ONT analysis, metadata only needs to match the sample columns in the beta matrix. `Sample_Name` must match `colnames(beta_all)`.

```{r create-meta, eval = FALSE}
sample_ids <- colnames(beta_all)

sample_group <- ifelse(
  grepl("-cfDNA-", sample_ids, ignore.case = TRUE),
  "cfDNA",
  ifelse(
    grepl("-WBC-", sample_ids, ignore.case = TRUE),
    "WBC",
    sub("-Z[0-9A-Za-z]+$", "",
        sub("^GSM[0-9]+_", "", sample_ids))
  )
)
meta_region <- data.frame(
  Sample_Name = colnames(beta_all),
  Sample_Group = sample_group,
  stringsAsFactors = FALSE
)

meta_region
```

---

## Phase 4: Create an ImprintomeSet

Region-level sequencing objects should explicitly set the correct `genome` (`"hg19"` or `"hg38"`) and a sequencing assay label such as `"WGBS"` or `"ONT"`.

```{r create-imprintomeset, eval = FALSE}
probesets <- readRDS(system.file(
  "extdata",
  "probesets_hg19.rds",
  package = "imprintomeR"
))

x <- ImprintomeSet(
  beta = beta_all,
  meta = meta_region,
  probeset = probesets[["Rosenski_region"]],
  genome = "hg19",
  assay = "WGBS"
)

x
```

---

## Phase 5: Run Region-Level Imprinting Analysis

`runImprintome()` uses the maternal and paternal `ORIGIN` annotations in `Rosenski_region` to calculate paternal/maternal medians, IDS, Angle, and imprinting status.

```{r run-imprintome, eval = FALSE}
x <- runImprintome(
  x,
  probeset = "Rosenski_region",
  ids_cutoff = 0.2
)

status <- results(x)[["AnalyzeImprintStatus.Rosenski_region"]]
status[, c(
  "Sample_Name",
  "paternal_median",
  "maternal_median",
  "IDS",
  "Angle",
  "Mechanism",
  "Status"
)]
```

Interpretation should remain cautious because region-level WGBS/ONT beta values are not allele-specific. IDS and Angle summarize directional deviation from balanced methylation across maternal-origin and paternal-origin regions.

---

## Phase 6: Region-Safe Default Visualizations

For `probeset = "Rosenski_region"` or `probeset = "chr11p15_region"`, the default visualization set is intentionally restricted to plot types that work with region-level beta rows:

```text
polar
beeswarm_origin
heatmap_by_gene
radar
```

```{r run-region-visualizations, eval = FALSE}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  result_name = "AnalyzeImprintStatus.Rosenski_region",
  plot_types = "default",
  save_plots = FALSE,
  store_plots = TRUE
)

attr(x, "visualization_manifest")
names(plots(x))
```

### Individual Plot Examples

```{r plot-polar, eval = FALSE, fig.width = 8, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  result_name = "AnalyzeImprintStatus.Rosenski_region",
  plot_types = "polar",
  save_plots = FALSE,
  store_plots = TRUE
)
print(plots(x)[["polar.Rosenski_region"]])
```

```{r plot-beeswarm-origin, eval = FALSE, fig.width = 8, fig.height = 5}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  plot_types = "beeswarm_origin",
  save_plots = FALSE,
  store_plots = TRUE
)
print(plots(x)[["beeswarm_origin.Rosenski_region"]])
```

```{r plot-heatmap-gene, eval = FALSE, fig.width = 8, fig.height = 6}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  plot_types = "heatmap_by_gene",
  save_plots = FALSE,
  store_plots = TRUE
)
plots(x)[["heatmap_by_gene.Rosenski_region"]]
```

```{r plot-radar, eval = FALSE, fig.width = 8, fig.height = 8}
x <- runImprintomeVisualizations(
  x,
  probeset = "Rosenski_region",
  result_name = "AnalyzeImprintStatus.Rosenski_region",
  plot_types = "radar",
  save_plots = FALSE,
  store_plots = TRUE
)
print(plots(x)[[grep("^radar\\.", names(plots(x)), value = TRUE)[1]]])
```

Avoid using `plot_types = "all"` for region-level probesets unless you explicitly want to test non-region-aware plot paths. In particular, classic `beeswarm` and direct `rainfall` are probe-oriented and are not part of the region-safe default workflow.

---

## Phase 7: Export Results

After analysis and optional visualizations, export result tables and plot files.

```{r export-results, eval = FALSE}
manifest <- export(
  x,
  outdir = "Rosenski_region_imprintome_output",
  prefix = "Rosenski_iDMRs",
  save_plots = TRUE,
  plot_device = "pdf",
  overwrite = TRUE
)

manifest
```

The export will include available result tables such as:

```text
Rosenski_iDMRs_results_AnalyzeImprintStatus.Rosenski_region.tsv
Rosenski_iDMRs_imprintomeSet.rds
```

## Toy Example: Submit WGBS Summary and Run imprintomeR CLI

The following example summarizes WGBS `.beta` files over the refined Rosenski iDMR regions, submits the job to LSF, runs the imprintomeR CLI on the region-level beta table, and then exports a radar plot for one selected sample. The ONT bedMethyl path above produces the same `-B` input structure, usually with `--genome hg38`.

```{bash summarize-wgbs-regions, eval = FALSE}
# Summarize methylation over refined Rosenski iDMR regions
cmd="wgbstools convert inst/extdata/Rosenski_refined_iDMRs_hg19.bed --genome hg19 -L -o Rosenski_refined_iDMRs_hg19.withCpG.bed && wgbstools beta_to_table --betas *.beta \
--output Rosenski_iDMRs_mean_beta.hg19.tsv \
Rosenski_refined_iDMRs_hg19.withCpG.bed"

prefix=wgbstools_iDMRs_beta
bsub -q priority -J ${prefix} -M 50000 -oo ${prefix}.log -eo ${prefix}.errlog "$cmd" 2>&1 >/dev/null
```

```{bash run-wgbs-cli, eval = FALSE}
# Run the imprintomeR WGBS region workflow
Rscript /PATH/TO/imprintomeR/scripts/run_imprintomeR.R \
  -B Rosenski_iDMRs_mean_beta.hg19.tsv \
  -m GSE186458_meta.txt \
  -o wgbs_imprintome_results_mean \
  --prefix Rosenski_WGBS \
  --probeset Rosenski_region \
  --genome hg19 \
  --plot-types default \
  -v

# Focused hg38 chr11p15 region workflow
Rscript /PATH/TO/imprintomeR/scripts/run_imprintomeR.R \
  -B ONT_Rosenski_hg38_beta.tsv \
  -m ONT_meta.txt \
  -o chr11p15_region_results \
  --prefix chr11p15_region \
  --probeset chr11p15_region \
  --genome hg38 \
  --plot-types default \
  -v
```

```{r radar-specific-sample, eval = FALSE, fig.width = 8, fig.height = 8}
# Generate a radar plot for one specific sample
library(imprintomeR)

sample_id1 <- "GSM5652245_Pancreas-Acinar-Z0000043X"
x <- readRDS("wgbs_imprintome_results_mean/Rosenski_WGBS_imprintomeSet.rds")

x <- runImprintomeVisualizations(
  x,
  outdir = "wgbs_imprintome_results_mean",
  plot_types = "radar",
  probeset = "Rosenski_region",
  sample_id = sample_id1,
  prefix = "Rosenski_WGBS",
  store_plots = FALSE,
  save_plots = TRUE
)
```
---

## Notes and Limitations

- `Rosenski_region` and `chr11p15_region` are **region-level** signatures, not Illumina probe-level probesets.
- `ORIGIN` is derived from primary germline DMR evidence in Rosenski et al. Sheet S10.
- `Oocyte gDMR` is treated as maternal; `Sperm gDMR` is treated as paternal.
- Secondary-only or no-primary-germline regions have `ORIGIN = NA` and do not drive maternal/paternal scoring.
- Region-level WGBS/ONT beta values are not allele-specific; interpretation should be cautious and supported by orthogonal evidence when possible.
